---
title: "Fig. 1: Diversity patterns across one wet and dry season in two Amazon forest types"
author: "Jessica Finck"
date: "2026-02-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

This script analyses microbial community diversity – that is alpha diversity and beta diversity metrics – using the `phyloseq` and `vegan` packages. The starting point is a `phyloseq` object with your community data, or other types of compositional data (e.g. molecular data).

### 1 \| Pre-processing: Define `phyloseq` objects for bacteria and fungi

Prior to analysis, we have to create a `ps` object from our data. This will be used for most of the downstream analysis.

**Important!** The file 'rooted_tree.nwk' is generated via QIIME2 pre-processing workflow. The detailed script can be [**found here**](https://github.com/JFinck1994/general_scripts/blob/main/qiime2_amplicon-16S-PE-read-processing.sh)**.** The step `phyloseq::phy_tree(tree_file.nwk)` is optional.

```{r phyloseq, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Load dependencies
library(readxl) # import xlsx files
library(ape)    # import nwk files
library(dplyr)  # data wrangling

## Import data 
otu_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "asv") %>% as.data.frame()
tax_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "tax") %>% as.data.frame()
data_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_bac <- ape::read.tree("BAC_rooted_tree_dada2.nwk")

otu_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "asv") %>% as.data.frame()
tax_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "tax") %>% as.data.frame()
data_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_fun <- ape::read.tree("FUN_rooted_tree_dada2.nwk")

## Format variable classes
library(dplyr)  # data wrangling

data_bac$site <- as.factor(data_bac$site)
data_bac$subsite <- as.factor(data_bac$subsite)
data_bac$season <- as.factor(data_bac$season)

data_fun$site <- as.factor(data_fun$site)
data_fun$subsite <- as.factor(data_fun$subsite)
data_fun$season <- as.factor(data_fun$season)

## Adjust taxa names  
tax_bac$Kingdom = gsub("d__","",as.character(tax_bac$Kingdom))
tax_bac$Phylum = gsub("p__","",as.character(tax_bac$Phylum))
tax_bac$Class = gsub("c__","",as.character(tax_bac$Class))
tax_bac$Order = gsub("o__","",as.character(tax_bac$Order))
tax_bac$Family = gsub("f__","",as.character(tax_bac$Family))
tax_bac$Genus = gsub("g__","",as.character(tax_bac$Genus))
tax_bac$Species = gsub("s__","",as.character(tax_bac$Species))
otu_bac <- otu_bac %>% tibble::column_to_rownames("asv")
tax_bac <- tax_bac %>% tibble::column_to_rownames("asv")
data_bac <- data_bac %>% tibble::column_to_rownames("smp_id")

tax_fun$Kingdom = gsub("k__","",as.character(tax_fun$Kingdom))
tax_fun$Phylum = gsub("p__","",as.character(tax_fun$Phylum))
tax_fun$Class = gsub("c__","",as.character(tax_fun$Class))
tax_fun$Order = gsub("o__","",as.character(tax_fun$Order))
tax_fun$Family = gsub("f__","",as.character(tax_fun$Family))
tax_fun$Genus = gsub("g__","",as.character(tax_fun$Genus))
tax_fun$Species = gsub("s__","",as.character(tax_fun$Species))

otu_fun <- otu_fun %>% tibble::column_to_rownames("asv")
tax_fun <- tax_fun %>% tibble::column_to_rownames("asv")
data_fun <- data_fun %>% tibble::column_to_rownames("smp_id")

## Assemble ps objects 
library(phyloseq)

otu_bac <- as.matrix(otu_bac)
tax_bac <- as.matrix(tax_bac)
OTU_B = phyloseq::otu_table(otu_bac, taxa_are_rows = T)
TAX_B = phyloseq::tax_table(tax_bac)
samples_B = phyloseq::sample_data(data_bac)
tree_B = phyloseq::phy_tree(tree_bac)
ps.bac <- phyloseq::phyloseq(OTU_B, TAX_B, samples_B, tree_B)

otu_fun <- as.matrix(otu_fun)
tax_fun <- as.matrix(tax_fun)
OTU_F = phyloseq::otu_table(otu_fun, taxa_are_rows = T)
TAX_F = phyloseq::tax_table(tax_fun)
samples_F = phyloseq::sample_data(data_fun)
tree_F = phyloseq::phy_tree(tree_fun)
ps.fun <- phyloseq::phyloseq(OTU_F, TAX_F, samples_F, tree_F)

ps.bac
ps.fun

## Tidy ps objects w/ microeco
library(microeco)
library(file2meco)

meco.fun = phyloseq2meco(ps.fun)
meco.bac = phyloseq2meco(ps.bac)

meco.bac$tax_table %>% base::subset(Kingdom == "k__Archaea" | Kingdom == "k__Bacteria")
meco.fun$tax_table %>% base::subset(Kingdom == "k__Fungi")

meco.bac$filter_pollution(taxa = c("mitochondria", "chloroplast"))
meco.fun$filter_pollution(taxa = c("mitochondria", "chloroplast"))

meco.bac$tidy_dataset()                                                         
meco.fun$tidy_dataset()

## Convert back to ps
ps.meco.bac = meco2phyloseq(meco.bac)
ps.meco.fun = meco2phyloseq(meco.fun)

ps.bac
ps.meco.bac
ps.fun
ps.meco.fun

## Normalize
library(microbiomeMarker) # TSS normalization
library(dplyr)            # data wrangling

ps.bac.tss = ps.meco.bac %>% microbiomeMarker::normalize(method = "TSS")
ps.fun.tss = ps.meco.fun %>% microbiomeMarker::normalize(method = "TSS")
```

### 2 \| Alpha diversity analysis

This script compares the alpha diversity metrics between groups, and visualizes them as bar plots w/ letter display to show between-group differences. The corresponding excel file looks like this:

| sampleID | metric1 | metric(n) | variable1 | variable2 | variable(n) |
|----------|---------|-----------|-----------|-----------|-------------|
|          |         |           |           |           |             |
|          |         |           |           |           |             |
|          |         |           |           |           |             |

*Variables* are metadata (e.g. season, forest type, sampling site), and *metrics* are alpha diversity metrics (e.g. Shannon, Chao1, Richness). *SampleID* are the sample names that are used in the ASV feature table.

```{r alphadiv1, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Import data
library(readxl)
bac_adiv <- read_excel("BAC_silva138.2.xlsx", sheet = "meta_div") %>% 
  as.data.frame() %>% 
  mutate(subsite = recode(subsite, "terrace" = "TE", "plateau" = "PL", 
                          "campinarana" = "C1","campina" = "C2")) 

fun_adiv <- read_excel("FUN_UNITE10.xlsx", sheet = "meta_div") %>% 
  as.data.frame() %>% 
  mutate(subsite = recode(subsite, "terrace" = "TE", "plateau" = "PL", 
                          "campinarana" = "C1","campina" = "C2")) 

## Group all possible combinations 
bac_adiv$Group <- with(bac_adiv, interaction(subsite, season, sep = "_"))
fun_adiv$Group <- with(fun_adiv, interaction(subsite, season, sep = "_"))

## Stats as letter display
anova_bac.shan <- aov(div_shan ~ Group, data = bac_adiv) # ANOVA
anova_fun.shan <- aov(div_shan ~ Group, data = fun_adiv) # ANOVA

tukey_bac.shan <- TukeyHSD(anova_bac.shan) # Tukey HSD Posthoc
tukey_fun.shan <- TukeyHSD(anova_fun.shan) # Tukey HSD Posthoc

library(multcompView)
cld_bac.shan <- multcompLetters4(anova_bac.shan, tukey_bac.shan) # letter display
cld_fun.shan <- multcompLetters4(anova_fun.shan, tukey_fun.shan) # letter display

cld_bac.shan <- data.frame(Group = names(cld_bac.shan$Group$Letters), 
                           Letters = cld_bac.shan$Group$Letters)
cld_fun.shan <- data.frame(Group = names(cld_fun.shan$Group$Letters), 
                           Letters = cld_fun.shan$Group$Letters)

letters_bac.shan <- tidyr::separate(cld_bac.shan, Group, 
                                    into = c("subsite", "season"), sep = "_")
letters_fun.shan <- tidyr::separate(cld_fun.shan, Group, 
                                    into = c("subsite", "season"), sep = "_")

letters_bac.shan$site_season <- rownames(letters_bac.shan)
letters_fun.shan$site_season <- rownames(letters_fun.shan)

## Plot alpha div bar plot 
library(dplyr)
library(ggpubr)

bac_adiv %>%
  mutate(Group = factor(Group, levels = c("PL_dry", "PL_wet", 
                                          "TE_dry", "TE_wet",
                                          "C1_dry", "C1_wet", 
                                          "C2_dry", "C2_wet"))) %>%  
  ggbarplot("Group","div_shan", fill = "Group", color = "Group", 
            palette = c("lightgreen","seagreen3", "lightgreen","seagreen3",
                        "bisque2", "bisque4","bisque2", "bisque4"),       
            position = position_dodge(0.8), 
            add = "mean_sd", error.plot = "upper_errorbar", width = 0.7) +
  geom_text(data = distinct(letters_bac.shan, site_season, Letters), 
            aes(x = site_season, y = max(bac_adiv$div_shan) + 0.2, 
                label = Letters), position = position_dodge(width = 0.8)) + 
  labs(x=NULL, y="Shannon index") +
  theme_pubr() + theme(panel.grid = element_blank(), legend.position = "top") +
  ggpubr::theme_pubr() + coord_flip() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "bottom", legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  # Add median as square
  stat_summary(aes(group = interaction(subsite, season)), 
               fun = median, geom = "point", shape = 0, size = 3, fill = "black",
               position = position_dodge(0.8)) +
  scale_x_discrete(labels = c("PL_dry" = "PL-dry","PL_wet" = "PL-wet",
                              "TE_dry" = "TE-dry","TE_wet" = "TE-wet",
                              "C1_dry" = "C1-dry","C1_wet" = "C1-wet",
                              "C2_dry" = "C2-dry","C2_wet" = "C2-wet"))

fun_adiv %>%
  mutate(Group = factor(Group, levels = c("PL_dry", "PL_wet", "TE_dry", "TE_wet",
                                          "C1_dry", "C1_wet", "C2_dry", "C2_wet"))) %>%  
  ggbarplot("Group","div_shan", fill = "Group", color = "Group", 
            palette = c("lightgreen","seagreen3","lightgreen","seagreen3",
                        "bisque2", "bisque4","bisque2", "bisque4"), 
            position = position_dodge(0.8), 
            add = "mean_sd", error.plot = "upper_errorbar", width = 0.7) +
  geom_text(data = distinct(letters_bac.shan, site_season, Letters), 
            aes(x = site_season, y = max(fun_adiv$div_shan) + 0.2, 
                label = Letters), position = position_dodge(width = 0.8)) + 
  labs(x=NULL, y="Shannon index") +
  theme_pubr() + theme(panel.grid = element_blank(), legend.position = "top") +
  ggpubr::theme_pubr() + coord_flip() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "bottom", legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  # Add median as square
  stat_summary(aes(group = interaction(subsite, season)), 
               fun = median, geom = "point", shape = 0, size = 3, fill = "black",
               position = position_dodge(0.8)) +
  scale_x_discrete(labels = c("PL_dry" = "PL-dry","PL_wet" = "PL-wet",
                              "TE_dry" = "TE-dry","TE_wet" = "TE-wet",
                              "C1_dry" = "C1-dry","C1_wet" = "C1-wet",
                              "C2_dry" = "C2-dry","C2_wet" = "C2-wet"))
```
