---
title: "Fig. 1: Diversity patterns across one wet and dry season in two Amazon forest types"
author: "Jessica Finck"
date: "2026-02-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

This script analyses microbial community diversity – that is alpha diversity and beta diversity metrics – using the `phyloseq` and `vegan` packages. The starting point is a `phyloseq` object with your community data, or other types of compositional data (e.g. molecular data).

### 1 \| Pre-processing: Define `phyloseq` objects for bacteria and fungi

Prior to analysis, we have to create a `ps` object from our data. This will be used for most of the downstream analysis.

**Important!** The file 'rooted_tree.nwk' is generated via QIIME2 pre-processing workflow. The detailed script can be [**found here**](https://github.com/JFinck1994/general_scripts/blob/main/qiime2_amplicon-16S-PE-read-processing.sh)**.** The step `phyloseq::phy_tree(tree_file.nwk)` is optional.

```{r phyloseq, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Load dependencies
library(readxl) # import xlsx files
library(ape)    # import nwk files
library(dplyr)  # data wrangling

## Import data 
otu_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "asv") %>% as.data.frame()
tax_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "tax") %>% as.data.frame()
data_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_bac <- ape::read.tree("BAC_rooted_tree_dada2.nwk")

otu_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "asv") %>% as.data.frame()
tax_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "tax") %>% as.data.frame()
data_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_fun <- ape::read.tree("FUN_rooted_tree_dada2.nwk")

## Format variable classes
library(dplyr)  # data wrangling

data_bac$site <- as.factor(data_bac$site)
data_bac$subsite <- as.factor(data_bac$subsite)
data_bac$season <- as.factor(data_bac$season)

data_fun$site <- as.factor(data_fun$site)
data_fun$subsite <- as.factor(data_fun$subsite)
data_fun$season <- as.factor(data_fun$season)

## Adjust taxa names  
tax_bac$Kingdom = gsub("d__","",as.character(tax_bac$Kingdom))
tax_bac$Phylum = gsub("p__","",as.character(tax_bac$Phylum))
tax_bac$Class = gsub("c__","",as.character(tax_bac$Class))
tax_bac$Order = gsub("o__","",as.character(tax_bac$Order))
tax_bac$Family = gsub("f__","",as.character(tax_bac$Family))
tax_bac$Genus = gsub("g__","",as.character(tax_bac$Genus))
tax_bac$Species = gsub("s__","",as.character(tax_bac$Species))
otu_bac <- otu_bac %>% tibble::column_to_rownames("asv")
tax_bac <- tax_bac %>% tibble::column_to_rownames("asv")
data_bac <- data_bac %>% tibble::column_to_rownames("smp_id")

tax_fun$Kingdom = gsub("k__","",as.character(tax_fun$Kingdom))
tax_fun$Phylum = gsub("p__","",as.character(tax_fun$Phylum))
tax_fun$Class = gsub("c__","",as.character(tax_fun$Class))
tax_fun$Order = gsub("o__","",as.character(tax_fun$Order))
tax_fun$Family = gsub("f__","",as.character(tax_fun$Family))
tax_fun$Genus = gsub("g__","",as.character(tax_fun$Genus))
tax_fun$Species = gsub("s__","",as.character(tax_fun$Species))

otu_fun <- otu_fun %>% tibble::column_to_rownames("asv")
tax_fun <- tax_fun %>% tibble::column_to_rownames("asv")
data_fun <- data_fun %>% tibble::column_to_rownames("smp_id")

## Assemble ps objects 
library(phyloseq)

otu_bac <- as.matrix(otu_bac)
tax_bac <- as.matrix(tax_bac)
OTU_B = phyloseq::otu_table(otu_bac, taxa_are_rows = T)
TAX_B = phyloseq::tax_table(tax_bac)
samples_B = phyloseq::sample_data(data_bac)
tree_B = phyloseq::phy_tree(tree_bac)
ps.bac <- phyloseq::phyloseq(OTU_B, TAX_B, samples_B, tree_B)

otu_fun <- as.matrix(otu_fun)
tax_fun <- as.matrix(tax_fun)
OTU_F = phyloseq::otu_table(otu_fun, taxa_are_rows = T)
TAX_F = phyloseq::tax_table(tax_fun)
samples_F = phyloseq::sample_data(data_fun)
tree_F = phyloseq::phy_tree(tree_fun)
ps.fun <- phyloseq::phyloseq(OTU_F, TAX_F, samples_F, tree_F)

ps.bac
ps.fun

## Tidy ps objects w/ microeco
library(microeco)
library(file2meco)

meco.fun = phyloseq2meco(ps.fun)
meco.bac = phyloseq2meco(ps.bac)

meco.bac$tax_table %>% base::subset(Kingdom == "k__Archaea" | Kingdom == "k__Bacteria")
meco.fun$tax_table %>% base::subset(Kingdom == "k__Fungi")

meco.bac$filter_pollution(taxa = c("mitochondria", "chloroplast"))
meco.fun$filter_pollution(taxa = c("mitochondria", "chloroplast"))

meco.bac$tidy_dataset()                                                         
meco.fun$tidy_dataset()

## Convert back to ps
ps.meco.bac = meco2phyloseq(meco.bac)
ps.meco.fun = meco2phyloseq(meco.fun)

ps.bac
ps.meco.bac
ps.fun
ps.meco.fun

## Normalize
library(microbiomeMarker) # TSS normalization
library(dplyr)            # data wrangling

ps.bac.tss = ps.meco.bac %>% microbiomeMarker::normalize(method = "TSS")
ps.fun.tss = ps.meco.fun %>% microbiomeMarker::normalize(method = "TSS")
```

### 2 \| Alpha diversity analysis (see other Rmd)

### 3 \| Beta diversity analsis

#### 3.1 \| Prerequisite steps

Initially, we filtered out the sample `DSTB` as stated and discussed in the manuscript (see 1 \| Pre-processing). We then extract our feature and sample data, and introduce a grouping variable for later plotting based on grouping (e.g. forest type, season). Lastly, we define a common ordination plot theme for consistency.

```{r rda1, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Load dependencies
library(dplyr)
library(tidyr)
library(phyloseq)
library(vegan)

## Extract & transform data for beta diversity ordination
otu.bac = ps.meco.bac %>% phyloseq::otu_table() %>% 
  t() %>% decostand(method = "hellinger") %>% as.data.frame()

otu.fun = ps.meco.fun %>% phyloseq::otu_table() %>% 
  t() %>% decostand(method = "hellinger") %>% as.data.frame()

data.bac = data.frame(phyloseq::sample_data(ps.meco.bac))
data.fun = data.frame(phyloseq::sample_data(ps.meco.fun))

## Make grouping variable
data.bac$Group <- paste(data.bac$site, data.bac$season, sep = "_")
data.bac$Group <- factor(data.bac$Group, levels = c(
  "TF_dry", "TF_wet", "WSE_dry", "WSE_wet"))

data.fun$Group <- paste(data.fun$site, data.bac$season, sep = "_")
data.fun$Group <- factor(data.fun$Group, levels = c(
  "TF_dry", "TF_wet", "WSE_dry", "WSE_wet"))

## PCOA / RDA theme 
theme_coord <- function (base_size = 12, base_family = "", border = FALSE, margin = TRUE, 
                         legend = c("bottom"), x.text.angle = 0) 
{
  half_line <- base_size/2
  if (!is.numeric(legend)) 
    legend <- match.arg(legend)
  if (x.text.angle > 5) 
    xhjust <- 1
  else xhjust <- NULL
  if (margin) 
    plot.margin <- margin(half_line, half_line, half_line, half_line)
  else plot.margin <- unit(c(0.5, 0.3, 0.3, 0.3), "mm")
  .theme <- 
    theme_bw(base_size = base_size, base_family = base_family) %+replace% 
    theme(panel.border = element_rect(fill = NA, colour = "black", size = 0.7),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(), 
          axis.line = element_blank(), 
          axis.text = element_text(color = "black"), 
          legend.key = element_blank(), 
          strip.background = element_rect(fill = "#F2F2F2", colour = "black", size = 0.7), 
          plot.margin = plot.margin, 
          legend.position = legend, complete = TRUE,
          plot.title = element_text(hjust = 0.5, face = "bold", size = 16, vjust = 2))
  if (x.text.angle != 0) 
    .theme <- .theme + theme(axis.text.x = element_text(angle = x.text.angle, 
                                                        hjust = xhjust))
  .theme
}
```

#### 3.2 \| Permutational analysis of variance (PERMANOVA) w/ `vegan`

While this is a separate table in the respective manuscript (supplementary Table S2), we include PERMANOVA in this script as it is a vital part of microbial community diversity analysis. Downstream, we use the PERMANOVA results to directly plot with our ordinations. So, to show how this is done, we include PERMANOVA as part of this code script. Since we use hellinger-transformed counts for the later RDA, we use the same distance type (=*euclidean*) for PERMANOVA.

```{r perm1, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Calc distance 
library(phyloseq)
euc_dist.ps.bac <- vegan::vegdist(otu.bac, method = "euclidean")
euc_dist.ps.fun <- vegan::vegdist(otu.fun, method = "euclidean")

## Check interaction between predictors
betadisper(euc_dist.ps.bac, group = interaction(data.bac$season, data.bac$site))
betadisper(euc_dist.ps.fun, group = interaction(data.fun$season, data.fun$site))
```

We then sequentially fit all variables and identify which have significant impacts on microbial community composition. For later RDA we only keep the significant-effect variables as constraints.

```{r perm2, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## PERMANOVA: sequentially-fitting variables
perm.bac = vegan::adonis2(euc_dist.ps.bac ~ season * site + subsite, 
                          data = data.bac, permutations = 999, by = "terms")
perm.fun = vegan::adonis2(euc_dist.ps.fun ~ season * site + subsite, 
                          data = data.fun, permutations = 999, by = "terms")

perm.bac
perm.fun

## PERMANOVA: globally (same formula)
perm.bac.global = vegan::adonis2(euc_dist.ps.bac ~ season * site + subsite, 
                                 data = data.bac, permutations = 999)
perm.fun.global = vegan::adonis2(euc_dist.ps.fun ~ season * site + subsite, 
                                 data = data.fun, permutations = 999)
perm.bac.global
perm.fun.global

## Define PERMANOVA labels for ordination plots
perm.R2_bac <- round(perm.bac.global$R2[1], 3)
perm.R2_fun <- round(perm.fun.global$R2[1], 3)

perm.p_bac  <- round(perm.bac.global$`Pr(>F)`[1], 3)
perm.p_fun  <- round(perm.fun.global$`Pr(>F)`[1], 3)

perm.label_bac <- paste0("PERMANOVA:\nR^2 = ", perm.R2_bac, "\np = ", perm.p_bac)
perm.label_fun <- paste0("PERMANOVA:\nRB2 = ", perm.R2_fun, "\np = ", perm.p_fun)
```

Lastly, we calculate the %Sum of Squares for later reporting in the manuscript. For this purpose, we defined a customary function to directly apply to our PERMANOVA results and retrieve %SS values.

```{r perm3, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Define function to calc %SS
get_permanova_stats <- function(adonis_obj) {
  aov_tab <- adonis_obj
  # Extract SS including residuals for %SS
  total_SS <- sum(aov_tab$SumOfSqs, na.rm = TRUE)
  # Remove residuals for table display
  aov_tab <- aov_tab[!rownames(aov_tab) %in% c("Residual", "Total"), ]
  Fval <- aov_tab$F
  pval <- aov_tab$`Pr(>F)`
  perc_SS <- round((aov_tab$SumOfSqs / total_SS) * 100, 3)
  data.frame(
    Term = rownames(aov_tab),
    F_value = round(Fval, 3),
    p_value = round(pval, 3),
    Percent_SS = perc_SS,
    row.names = NULL
  )
}

## Calc %SS
print(get_permanova_stats(perm.bac))
print(get_permanova_stats(perm.fun))
```

#### 3.3 \| Redundancy analysis (RDA) w/ `vegan`

After testing which variables (e.g. season, forest type/site, sampling location/subsite) have a significant impact on community composition using PERMANOVA, we use these variables to constrain the RDA ordination.

```{r rda2, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Computed mixed models based on PERMANOVA results
rda_model.bac = vegan::rda(otu.bac ~ season * site + subsite, data = data.bac)
rda_model.fun = vegan::rda(otu.fun ~ season * site + subsite, data = data.fun)

## Calculate RDA axis scores
eigenvalues.bac <- summary(rda_model.bac)$concont$importance["Eigenvalue", ]
eigenvalues.fun <- summary(rda_model.fun)$concont$importance["Eigenvalue", ]
variance.bac <- eigenvalues.bac / sum(eigenvalues.bac)
variance.fun <- eigenvalues.fun / sum(eigenvalues.fun)

variance.bac = variance.bac %>% as.list()
variance.fun = variance.fun %>% as.list()

## RDA axis scores
variance.bac
variance.fun
```

Now, we can compute ordination based on the RDA models.

**Important!** RDA itself is – basically – a constrained Principal Component Analysis, operating in a mathematically Euclidean space. Hence, it requires data to be normalized and transformed accordingly. If we want to use RDA, we thus have to apply either CLR transformation, rCLR transformation, or hellinger transformation, the latter being really common. Using rCLR transformation will evaluate robust Aitchison (rAitchison) distance instead. The alternative to PCA/RDA that operate in an Euclidean space is Principal Coordinate Analysis (PCoA), which is based on e.g. Bray-Curtis dissimilarity, and is commonly used in ecological studies.

```{r rda3, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Extract ordination scores
scores.bac <- scores(rda_model.bac, display = "sites")
scores.fun <- scores(rda_model.fun, display = "sites")

## Compute RDA ordination
ordination.rda_bac <- data.frame(scores.bac, 
                                 season = data.bac$season, 
                                 site = data.bac$site, 
                                 subsite = data.bac$subsite)
ordination.rda_fun <- data.frame(scores.fun, 
                                 season = data.fun$season, 
                                 site = data.fun$site, 
                                 subsite = data.fun$subsite)

ordination.rda_bac$Group <- data.bac$Group
ordination.rda_fun$Group <- data.fun$Group
ordination.rda_bac$smpID <- rownames(data.bac)
ordination.rda_fun$smpID <- rownames(data.fun)

## Plot RDA 
library(ggplot2) 

ggplot(ordination.rda_bac, aes(x=RDA1, y=RDA2, color = Group, shape = subsite)) + 
  geom_point(size = 5, alpha = 0.5) + theme_coord() + 
  scale_color_manual(values = c("TF_dry" = "lightgreen", "TF_wet" = "forestgreen", 
                                "WSE_dry" = "bisque", "WSE_wet" = "bisque4")) +
  scale_fill_manual(values = c("TF_dry" = "lightgreen", "TF_wet" = "forestgreen", 
                               "WSE_dry" = "bisque", "WSE_wet" = "bisque4")) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") + 
  scale_shape_manual(values=c(18, 15, 17, 16)) +
  labs(title = "16S-ASVs", 
       x = paste0("RDA1 (", round(variance.bac$RDA1*100, 2), "%)"), 
       y = paste0("RDA2 (", round(variance.bac$RDA2*100, 2), "%)")) +
  annotate("text", 
           x = min(ordination.rda_bac$RDA1) * 1, 
           y = min(ordination.rda_bac$RDA2) * 0.8, 
           label = perm.label_bac, hjust = 0, size = 4) +
  scale_x_continuous(limits = c(-0.85, 0.55), 
                     labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  scale_y_continuous(limits = c(-0.55, 0.7), 
                     labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  geom_text(aes(label = smpID), check_overlap = TRUE) 
  
ggplot(ordination.rda_fun, aes(x=RDA1, y=RDA2, color = Group, shape = subsite)) + 
  geom_point(size = 5, alpha = 0.5) + theme_coord() + 
  scale_color_manual(values = c("TF_dry" = "lightgreen", "TF_wet" = "forestgreen", 
                                "WSE_dry" = "bisque", "WSE_wet" = "bisque4")) +
  scale_fill_manual(values = c("TF_dry" = "lightgreen", "TF_wet" = "forestgreen", 
                               "WSE_dry" = "bisque", "WSE_wet" = "bisque4")) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") + 
  scale_shape_manual(values=c(18, 15, 17, 16)) +
  labs(title = "ITS-ASVs", 
       x = paste0("RDA1 (", round(variance.fun$RDA1*100, 2), "%)"), 
       y = paste0("RDA2 (", round(variance.fun$RDA2*100, 2), "%)")) +
  annotate("text", 
           x = min(ordination.rda_fun$RDA1) * 1, 
           y = min(ordination.rda_fun$RDA2) * 0.8, 
           label = perm.label_fun, hjust = 0, size = 4) +
  scale_x_continuous(limits = c(-0.52,0.7), 
                     labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  scale_y_continuous(limits = c(-0.85,0.5), 
                     labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  geom_text(aes(label = smpID), check_overlap = TRUE) 
```

#### 3.4 \| Principal coordinate analysis (PCoA) w/ `vegan`

While we used RDA for our manuscript, PCoA can also be used, and is commonly used, to visualize beta diversity of microbial communities. Hence, we provide a comprehensive script to run PCoA too. Here, we chose Bray-Curtis dissimilarity matrix, which is calculated based on normalized (total sum scaling) counts.

```{r pcoa, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(phyloseq)
library(vegan)
library(ggplot2)
library(ggrepel)

## Calculate distance matrix
dist_bac <- phyloseq::distance(ps.bac.tss, method = "bray")
dist_fun <- phyloseq::distance(ps.fun.tss, method = "bray")

## Calculate ordination based on dist
pcoa_bac <- ordinate(ps.bac.tss, method = "PCoA", distance = dist_bac)
pcoa_fun <- ordinate(ps.fun.tss, method = "PCoA", distance = dist_fun)

## Plot PCoA
plot_ordination(ps.bac.tss, pcoa_bac, color = "Group", shape = "subsite") +
  geom_point(aes(size = 5, alpha = 0.5)) + 
  theme_coord() + 
  labs(title = "16S-ASVs") + 
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") +
  scale_shape_manual(values = c(18, 15, 17, 16)) +
  annotate("text", 
           x = min(pcoa_bac$Axis.1) * 0.75, 
           y = min(pcoa_bac$Axis.2) * 0.325, 
           label = perm.label_bac, hjust = 0, size = 4)

plot_ordination(ps.fun.tss, pcoa_fun, color = "Group", shape = "subsite") +
  geom_point(aes(size = 5, alpha = 0.5)) + 
  theme_coord() + 
  labs(title = "ITS-ASVs") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") +
  scale_shape_manual(values = c(18, 15, 17, 16)) +
  annotate("text", 
           x = min(pcoa_fun$Axis.1) * 0.6, 
           y = min(pcoa_fun$Axis.2) * 0.4, 
           label = perm.label_fun, hjust = 0, size = 4) 
```

#### 3.5 \| Beta diversity distance

After importing our data, we'll apply the filter to analyse only putative SOM-cycling taxa, creating a joint bacteria-fungi df for subsequent plotting with `ggplot2`.

```{r betadiv1, message=FALSE, warning=FALSE, results='hide', tidy=FALSE}
library(vegan)       # for distance matrix calculation
library(ggplot2)     # for plotting
library(dplyr)       # for data wrangling
library(tidyr)       # for data wrangling
library(ggsignif)    # optional, for significance marks

## Already exists from step (1 | Pre-processing) 
otu.bac # hellinger-transformed counts
otu.fun # hellinger-transformed counts 
dist_bac.hell <- vegdist(otu.bac, method = "euclidean")
dist_fun.hell <- vegdist(otu.fun, method = "euclidean")

## Create grouping variables
data.bac$Group <- paste(data.bac$site, data.bac$season, sep = "_")
data.bac$Group <- factor(data.bac$Group, levels = c("TF_dry", "TF_wet", "WSE_dry", "WSE_wet"))

data.fun$Group <- paste(data.fun$site, data.bac$season, sep = "_")
data.fun$Group <- factor(data.fun$Group, levels = c("TF_dry", "TF_wet", "WSE_dry", "WSE_wet"))

data.bac$Group2 <- paste(data.bac$subsite, data.bac$season, sep = "_")
data.bac$Group2 <- factor(data.bac$Group2, levels = c(
  "plateau_dry","campinarana_dry","campina_dry","terrace_dry",
  "campina_wet","campinarana_wet","plateau_wet","terrace_wet"))

data.fun$Group2 <- paste(data.fun$subsite, data.bac$season, sep = "_")
data.fun$Group2 <- factor(data.fun$Group2, levels = c(
  "plateau_dry","campinarana_dry","campina_dry","terrace_dry",
  "campina_wet","campinarana_wet","plateau_wet","terrace_wet"))

## Create dist_df long-format 
dist_bac.df <- as.data.frame(as.table(as.matrix(dist_bac.hell))) %>%
  rename(sample1 = Var1, sample2 = Var2, distance = Freq) %>%
  filter(as.character(sample1) < as.character(sample2)) %>%  # upper triangle only
  mutate(group1 = data.bac[sample1, "Group"], group2 = data.bac[sample2, "Group"]) %>%
  # keep only within-group distances
  filter(group1 == group2) %>%
  select(distance, group = group1)

dist_fun.df <- as.data.frame(as.table(as.matrix(dist_fun.hell))) %>%
  rename(sample1 = Var1, sample2 = Var2, distance = Freq) %>%
  filter(as.character(sample1) < as.character(sample2)) %>%
  mutate(group1 = data.fun[sample1, "Group"], group2 = data.fun[sample2, "Group"]) %>%
  filter(group1 == group2) %>%
  select(distance, group = group1)

dist_bac.df2 <- as.data.frame(as.table(as.matrix(dist_bac.hell))) %>%
  rename(sample1 = Var1, sample2 = Var2, distance = Freq) %>%
  filter(as.character(sample1) < as.character(sample2)) %>%
  mutate(group1 = data.bac[sample1, "Group2"], group2 = data.bac[sample2, "Group2"]) %>%
  filter(group1 == group2) %>%
  select(distance, group = group1)

dist_fun.df2 <- as.data.frame(as.table(as.matrix(dist_fun.hell))) %>%
  rename(sample1 = Var1, sample2 = Var2, distance = Freq) %>%
  filter(as.character(sample1) < as.character(sample2)) %>%
  mutate(group1 = data.fun[sample1, "Group2"], group2 = data.fun[sample2, "Group2"]) %>%
  filter(group1 == group2) %>%
  select(distance, group = group1)

## Calculate stats letters
library(multcompView) 
anova_bac <- aov(distance ~ group, data = dist_bac.df)
tukey_bac <- TukeyHSD(anova_bac)
letters_bac <- multcompLetters4(anova_bac, tukey_bac)
anova_fun <- aov(distance ~ group, data = dist_fun.df)
tukey_fun <- TukeyHSD(anova_fun)
letters_fun <- multcompLetters4(anova_fun, tukey_fun)
letters_bac$group
letters_fun$group

## Set plotting order & rename to desired labels 
dist_bac.df$group <- factor(dist_bac.df$group, 
                            levels = c("TF_dry","TF_wet","WSE_dry","WSE_wet"),
                            labels = c("TF-dry","TF-wet","WS-dry","WS-wet"))
dist_fun.df$group <- factor(dist_fun.df$group, 
                            levels = c("TF_dry","TF_wet","WSE_dry","WSE_wet"),
                            labels = c("TF-dry","TF-wet","WS-dry","WS-wet"))

dist_bac.df2$group <- factor(dist_bac.df2$group, 
                             levels = c("plateau_dry","plateau_wet",
                                        "terrace_dry","terrace_wet",
                                        "campinarana_dry","campinarana_wet",
                                        "campina_dry","campina_wet"),
                             labels = c("PL-dry","PL-wet","TE-dry","TE-wet",
                                        "C1-dry","C1-wet","C2-dry","C2-wet"))
dist_fun.df2$group <- factor(dist_fun.df2$group, 
                             levels = c("plateau_dry","plateau_wet",
                                        "terrace_dry","terrace_wet",
                                        "campinarana_dry","campinarana_wet",
                                        "campina_dry","campina_wet"),
                             labels = c("PL-dry","PL-wet","TE-dry","TE-wet",
                                        "C1-dry","C1-wet","C2-dry","C2-wet"))

## Plot as bxp (resolution = forest type)
library(ggplot2)
library(scales)

dist_bac.df %>%
  ggplot(aes(x = group, y = distance, fill = group)) + 
  geom_boxplot() + 
  ggpubr::theme_pubr() + 
  scale_fill_manual(values = c(
    "TF-dry" = "lightgreen","WS-dry"= "bisque",
    "WS-wet"= "bisque4","TF-wet"= "forestgreen")) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  ggpubr::theme_pubr() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 0), 
        legend.position = "bottom", legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  labs(x=NULL, y="Hellinger distance") +
  geom_jitter(aes(shape = subsite), width = 0.2, size = 2.5, color = "black", alpha = 0.5) +
    scale_shape_manual(values = c("PL" = 17,"TE" = 16,"C1" = 15,"C2" = 18)) +
  dist_bac.df2 %>%
  rename(group2 = group) %>%
  mutate(group = case_when(group2 %in% c("PL-wet", "TE-wet") ~ "TF-wet", 
                           group2 %in% c("PL-dry", "TE-dry") ~ "TF-dry",
                           group2 %in% c("C1-dry", "C2-dry") ~ "WS-dry", 
                           group2 %in% c("C1-wet", "C2-wet") ~ "WS-wet", 
                           TRUE ~ as.character(group2)),
         subsite = case_when(grepl("^PL", group2) ~ "PL",
                             grepl("^TE", group2) ~ "TE",
                             grepl("^C1", group2) ~ "C1",
                             grepl("^C2", group2) ~ "C2",TRUE ~ NA_character_)) 

dist_fun.df %>%
  ggplot(aes(x = group, y = distance, fill = group)) + 
  geom_boxplot() + 
  ggpubr::theme_pubr() + 
  scale_fill_manual(values = c(
    "TF-dry" = "lightgreen","WS-dry"= "bisque",
    "WS-wet"= "bisque4","TF-wet"= "forestgreen")) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  # Panel border black
  ggpubr::theme_pubr() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 0), 
        legend.position = "bottom", 
        legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  labs(x=NULL, y="Hellinger distance") +
  geom_jitter(aes(shape = subsite), width = 0.2, size = 2.5, color = "black", alpha = 0.5) +
  scale_shape_manual(values = c("PL" = 17,"TE" = 16,"C1" = 15,"C2" = 18)) +
  dist_fun.df2 %>%
  rename(group2 = group) %>%
  mutate(group = case_when(group2 %in% c("PL-wet", "TE-wet") ~ "TF-wet", 
                           group2 %in% c("PL-dry", "TE-dry") ~ "TF-dry",
                           group2 %in% c("C1-dry", "C2-dry") ~ "WS-dry", 
                           group2 %in% c("C1-wet", "C2-wet") ~ "WS-wet", 
                           TRUE ~ as.character(group2)),
         subsite = case_when(grepl("^PL", group2) ~ "PL",
                             grepl("^TE", group2) ~ "TE",
                             grepl("^C1", group2) ~ "C1",
                             grepl("^C2", group2) ~ "C2",TRUE ~ NA_character_))
```
