---
author: "Jessica Finck"
date: "2026-02-01"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

This script analyses microbial community diversity – that is alpha diversity and beta diversity metrics – using the `phyloseq` and `vegan` packages. The starting point is a `phyloseq` object with your community data, or other types of compositional data (e.g. molecular data).

### 1 \| Pre-processing: Define `phyloseq` objects for bacteria and fungi

Prior to analysis, we have to create a `ps` object from our data. This will be used for most of the downstream analysis.

**Important!** The file 'rooted_tree.nwk' is generated via QIIME2 pre-processing workflow. The detailed script can be [**found here**](https://github.com/JFinck1994/general_scripts/blob/main/qiime2_amplicon-16S-PE-read-processing.sh)**.** The step `phyloseq::phy_tree(tree_file.nwk)` is optional.

```{r phyloseq, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Load dependencies
library(readxl) # import xlsx files
library(ape)    # import nwk files
library(dplyr)  # data wrangling

## Import data 
otu_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "asv") %>% as.data.frame()
tax_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "tax") %>% as.data.frame()
data_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_bac <- ape::read.tree("BAC_rooted_tree_dada2.nwk")

otu_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "asv") %>% as.data.frame()
tax_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "tax") %>% as.data.frame()
data_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_fun <- ape::read.tree("FUN_rooted_tree_dada2.nwk")

## Format variable classes
library(dplyr)  # data wrangling

data_bac$site <- as.factor(data_bac$site)
data_bac$subsite <- as.factor(data_bac$subsite)
data_bac$season <- as.factor(data_bac$season)

data_fun$site <- as.factor(data_fun$site)
data_fun$subsite <- as.factor(data_fun$subsite)
data_fun$season <- as.factor(data_fun$season)

## Adjust taxa names  
tax_bac$Kingdom = gsub("d__","",as.character(tax_bac$Kingdom))
tax_bac$Phylum = gsub("p__","",as.character(tax_bac$Phylum))
tax_bac$Class = gsub("c__","",as.character(tax_bac$Class))
tax_bac$Order = gsub("o__","",as.character(tax_bac$Order))
tax_bac$Family = gsub("f__","",as.character(tax_bac$Family))
tax_bac$Genus = gsub("g__","",as.character(tax_bac$Genus))
tax_bac$Species = gsub("s__","",as.character(tax_bac$Species))
otu_bac <- otu_bac %>% tibble::column_to_rownames("asv")
tax_bac <- tax_bac %>% tibble::column_to_rownames("asv")
data_bac <- data_bac %>% tibble::column_to_rownames("smp_id")

tax_fun$Kingdom = gsub("k__","",as.character(tax_fun$Kingdom))
tax_fun$Phylum = gsub("p__","",as.character(tax_fun$Phylum))
tax_fun$Class = gsub("c__","",as.character(tax_fun$Class))
tax_fun$Order = gsub("o__","",as.character(tax_fun$Order))
tax_fun$Family = gsub("f__","",as.character(tax_fun$Family))
tax_fun$Genus = gsub("g__","",as.character(tax_fun$Genus))
tax_fun$Species = gsub("s__","",as.character(tax_fun$Species))

otu_fun <- otu_fun %>% tibble::column_to_rownames("asv")
tax_fun <- tax_fun %>% tibble::column_to_rownames("asv")
data_fun <- data_fun %>% tibble::column_to_rownames("smp_id")

## Assemble ps objects 
library(phyloseq)

otu_bac <- as.matrix(otu_bac)
tax_bac <- as.matrix(tax_bac)
OTU_B = phyloseq::otu_table(otu_bac, taxa_are_rows = T)
TAX_B = phyloseq::tax_table(tax_bac)
samples_B = phyloseq::sample_data(data_bac)
tree_B = phyloseq::phy_tree(tree_bac)
ps.bac <- phyloseq::phyloseq(OTU_B, TAX_B, samples_B, tree_B)

otu_fun <- as.matrix(otu_fun)
tax_fun <- as.matrix(tax_fun)
OTU_F = phyloseq::otu_table(otu_fun, taxa_are_rows = T)
TAX_F = phyloseq::tax_table(tax_fun)
samples_F = phyloseq::sample_data(data_fun)
tree_F = phyloseq::phy_tree(tree_fun)
ps.fun <- phyloseq::phyloseq(OTU_F, TAX_F, samples_F, tree_F)

ps.bac
ps.fun

## Tidy ps objects w/ microeco
library(microeco)
library(file2meco)

meco.fun = phyloseq2meco(ps.fun)
meco.bac = phyloseq2meco(ps.bac)

meco.bac$tax_table %>% base::subset(Kingdom == "k__Archaea" | Kingdom == "k__Bacteria")
meco.fun$tax_table %>% base::subset(Kingdom == "k__Fungi")

meco.bac$filter_pollution(taxa = c("mitochondria", "chloroplast"))
meco.fun$filter_pollution(taxa = c("mitochondria", "chloroplast"))

meco.bac$tidy_dataset()                                                         
meco.fun$tidy_dataset()

## Convert back to ps
ps.meco.bac = meco2phyloseq(meco.bac)
ps.meco.fun = meco2phyloseq(meco.fun)

ps.bac
ps.meco.bac
ps.fun
ps.meco.fun

## Remove DSTB
ps.meco.bac <- prune_samples(sample_names(ps.meco.bac) != "DSTB", ps.meco.bac)

## Normalize
library(microbiomeMarker) # TSS normalization
library(dplyr)            # data wrangling

ps.bac.tss = ps.meco.bac %>% microbiomeMarker::normalize(method = "TSS")
ps.fun.tss = ps.meco.fun %>% microbiomeMarker::normalize(method = "TSS")
```

### 2 \| Alpha diversity analysis (see other Rmd)

### 3 \| Beta diversity analsis (see other Rmd)

### 4 \| Other analyses

#### 4.1 \| Beta nearest taxon index (beta-NTI) w/ `picante` (see other Rmd)

#### 4.2 \| Beta diversity dispersion w/ `vegan`

Beta diversity dispersion – aka the homogeneity of multivariate dispersions – measures *how much species composition varies within a group*. Beta dispersion is calculated as the average distance from individual samples to their group centroid or median. It is a crucial tool in ecological and microbiome studies for comparing beta diversity variability between groups, but is not to be confused with distance variability in the data.

**Important!** We did not include this in the manuscript, but found it a useful tool to introduce under the umbrella of microbial diversity/community analysis regardless.

```{r betadisper1, message=FALSE, warning=FALSE, results='hide', tidy=FALSE}
## Extract & transform data for beta diversity ordination
library(phyloseq)
library(vegan)

otu.bac = ps.meco.bac %>% phyloseq::otu_table() %>% 
  t() %>% decostand(method = "hellinger") %>% as.data.frame()

otu.fun = ps.meco.fun %>% phyloseq::otu_table() %>% 
  t() %>% decostand(method = "hellinger") %>% as.data.frame()

data.bac = data.frame(phyloseq::sample_data(ps.meco.bac))
data.fun = data.frame(phyloseq::sample_data(ps.meco.fun))

## Define grouping column
data.bac2 = data.bac %>%
  mutate(site = recode(site, "WSE" = "WS")) %>%
  mutate(Group = paste(site, season, sep = "_"))
data.bac2$Group <- as.factor(data.bac2$Group)

data.fun2 = data.fun %>%
  mutate(site = recode(site, "WSE" = "WS")) %>%
  mutate(Group = paste(site, season, sep = "_"))
data.fun2$Group <- as.factor(data.fun2$Group)

## Calculate distance 
euc_dist.ps.bac <- vegan::vegdist(otu.bac, method = "euclidean")
euc_dist.ps.fun <- vegan::vegdist(otu.fun, method = "euclidean")

## Calculate beta dispersion
## - the distances these are based on are the same as used for PERMANOVA (see 3.2)
## - in this case: hellinger-based euclidean distances 
library(vegan)
bd_bac <- betadisper(euc_dist.ps.bac, group = data.bac2$Group)
bd_fun <- betadisper(euc_dist.ps.fun, group = data.fun2$Group)

## Extract distance to centroid 
bd_dist.bac <- data.frame(distance = bd_bac$distances, group = bd_bac$group)
bd_dist.fun <- data.frame(distance = bd_fun$distances, group = bd_fun$group)

## Run stats w/ ANOVA-TukeyHSD
anova_bd.bac <- aov(distance ~ group, data = bd_dist.bac)
anova_bd.fun <- aov(distance ~ group, data = bd_dist.fun)

perm_bd.bac  <- permutest(bd_bac, permutations = 999)
perm_bd.fun  <- permutest(bd_fun, permutations = 999)

tukey_bd.bac <- TukeyHSD(anova_bd.bac)
tukey_bd.fun <- TukeyHSD(anova_bd.fun)

## Extract stats as letter display
library(multcompView)
letters_bd.bac <- multcompLetters4(anova_bd.bac, tukey_bd.bac)
letters_bd.fun <- multcompLetters4(anova_bd.fun, tukey_bd.fun)

## Set plotting order
bd_dist.bac$group <- factor(bd_dist.bac$group, 
                            levels = c("TF_dry","TF_wet","WS_dry","WS_wet"),
                            labels = c("TF-dry","TF-wet","WS-dry","WS-wet"))
bd_dist.fun$group <- factor(bd_dist.fun$group, 
                            levels = c("TF_dry","TF_wet","WS_dry","WS_wet"),
                            labels = c("TF-dry","TF-wet","WS-dry","WS-wet"))

## Plot beta dispersion per group
library(dplyr)
library(ggplot2)
library(ggpubr)

bd_dist.bac %>%
  ggplot(aes(x = group, y = distance, fill = group)) + 
  geom_boxplot() + ggpubr::theme_pubr() + labs(x=NULL, y="Distance") +
  scale_fill_manual(values = c("TF-dry" = "lightgreen","WS-dry"= "bisque",
                               "WS-wet"= "bisque4","TF-wet"= "forestgreen")) +
    scale_y_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  ggpubr::theme_pubr() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "bottom", 
        legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  labs(y="Beta diversity dispersion")

bd_dist.fun %>%
  ggplot(aes(x = group, y = distance, fill = group)) + 
  geom_boxplot() + ggpubr::theme_pubr() + labs(x=NULL, y="Distance") +
  scale_fill_manual(values = c("TF-dry" = "lightgreen","WS-dry"= "bisque",
                               "WS-wet"= "bisque4","TF-wet"= "forestgreen")) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  ggpubr::theme_pubr() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "bottom", 
        legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  labs(y="Beta diversity dispersion")
```
