---
author: "Jessica Finck"
date: "2026-02-01"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

This script analyses microbial community diversity – that is alpha diversity and beta diversity metrics – using the `phyloseq` and `vegan` packages. The starting point is a `phyloseq` object with your community data, or other types of compositional data (e.g. molecular data).

### 1 \| Pre-processing: Define `phyloseq` objects for bacteria and fungi

Prior to analysis, we have to create a `ps` object from our data. This will be used for most of the downstream analysis.

**Important!** The file 'rooted_tree.nwk' is generated via QIIME2 pre-processing workflow. The detailed script can be [**found here**](https://github.com/JFinck1994/general_scripts/blob/main/qiime2_amplicon-16S-PE-read-processing.sh)**.** The step `phyloseq::phy_tree(tree_file.nwk)` is optional.

```{r phyloseq, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Load dependencies
library(readxl) # import xlsx files
library(ape)    # import nwk files
library(dplyr)  # data wrangling

## Import data 
otu_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "asv") %>% as.data.frame()
tax_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "tax") %>% as.data.frame()
data_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_bac <- ape::read.tree("BAC_rooted_tree_dada2.nwk")

otu_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "asv") %>% as.data.frame()
tax_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "tax") %>% as.data.frame()
data_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_fun <- ape::read.tree("FUN_rooted_tree_dada2.nwk")

## Format variable classes
library(dplyr)  # data wrangling

data_bac$site <- as.factor(data_bac$site)
data_bac$subsite <- as.factor(data_bac$subsite)
data_bac$season <- as.factor(data_bac$season)

data_fun$site <- as.factor(data_fun$site)
data_fun$subsite <- as.factor(data_fun$subsite)
data_fun$season <- as.factor(data_fun$season)

## Adjust taxa names  
tax_bac$Kingdom = gsub("d__","",as.character(tax_bac$Kingdom))
tax_bac$Phylum = gsub("p__","",as.character(tax_bac$Phylum))
tax_bac$Class = gsub("c__","",as.character(tax_bac$Class))
tax_bac$Order = gsub("o__","",as.character(tax_bac$Order))
tax_bac$Family = gsub("f__","",as.character(tax_bac$Family))
tax_bac$Genus = gsub("g__","",as.character(tax_bac$Genus))
tax_bac$Species = gsub("s__","",as.character(tax_bac$Species))
otu_bac <- otu_bac %>% tibble::column_to_rownames("asv")
tax_bac <- tax_bac %>% tibble::column_to_rownames("asv")
data_bac <- data_bac %>% tibble::column_to_rownames("smp_id")

tax_fun$Kingdom = gsub("k__","",as.character(tax_fun$Kingdom))
tax_fun$Phylum = gsub("p__","",as.character(tax_fun$Phylum))
tax_fun$Class = gsub("c__","",as.character(tax_fun$Class))
tax_fun$Order = gsub("o__","",as.character(tax_fun$Order))
tax_fun$Family = gsub("f__","",as.character(tax_fun$Family))
tax_fun$Genus = gsub("g__","",as.character(tax_fun$Genus))
tax_fun$Species = gsub("s__","",as.character(tax_fun$Species))

otu_fun <- otu_fun %>% tibble::column_to_rownames("asv")
tax_fun <- tax_fun %>% tibble::column_to_rownames("asv")
data_fun <- data_fun %>% tibble::column_to_rownames("smp_id")

## Assemble ps objects 
library(phyloseq)

otu_bac <- as.matrix(otu_bac)
tax_bac <- as.matrix(tax_bac)
OTU_B = phyloseq::otu_table(otu_bac, taxa_are_rows = T)
TAX_B = phyloseq::tax_table(tax_bac)
samples_B = phyloseq::sample_data(data_bac)
tree_B = phyloseq::phy_tree(tree_bac)
ps.bac <- phyloseq::phyloseq(OTU_B, TAX_B, samples_B, tree_B)

otu_fun <- as.matrix(otu_fun)
tax_fun <- as.matrix(tax_fun)
OTU_F = phyloseq::otu_table(otu_fun, taxa_are_rows = T)
TAX_F = phyloseq::tax_table(tax_fun)
samples_F = phyloseq::sample_data(data_fun)
tree_F = phyloseq::phy_tree(tree_fun)
ps.fun <- phyloseq::phyloseq(OTU_F, TAX_F, samples_F, tree_F)

ps.bac
ps.fun

## Tidy ps objects w/ microeco
library(microeco)
library(file2meco)

meco.fun = phyloseq2meco(ps.fun)
meco.bac = phyloseq2meco(ps.bac)

meco.bac$tax_table %>% base::subset(Kingdom == "k__Archaea" | Kingdom == "k__Bacteria")
meco.fun$tax_table %>% base::subset(Kingdom == "k__Fungi")

meco.bac$filter_pollution(taxa = c("mitochondria", "chloroplast"))
meco.fun$filter_pollution(taxa = c("mitochondria", "chloroplast"))

meco.bac$tidy_dataset()                                                         
meco.fun$tidy_dataset()

## Convert back to ps
ps.meco.bac = meco2phyloseq(meco.bac)
ps.meco.fun = meco2phyloseq(meco.fun)

ps.bac
ps.meco.bac
ps.fun
ps.meco.fun

## Normalize
library(microbiomeMarker) # TSS normalization
library(dplyr)            # data wrangling

ps.bac.tss = ps.meco.bac %>% microbiomeMarker::normalize(method = "TSS")
ps.fun.tss = ps.meco.fun %>% microbiomeMarker::normalize(method = "TSS")
```

### 2 \| Alpha diversity analysis

This script compares the alpha diversity metrics between groups, and visualizes them as bar plots w/ letter display to show between-group differences. The corresponding excel file looks like this:

| sampleID | metric1 | metric(n) | variable1 | variable2 | variable(n) |
|----------|---------|-----------|-----------|-----------|-------------|
|          |         |           |           |           |             |
|          |         |           |           |           |             |
|          |         |           |           |           |             |

*Variables* are metadata (e.g. season, forest type, sampling site), and *metrics* are alpha diversity metrics (e.g. Shannon, Chao1, Richness). *SampleID* are the sample names that are used in the ASV feature table.

```{r alphadiv1, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Import data
library(readxl)
bac_adiv <- read_excel("BAC_silva138.2.xlsx", sheet = "meta_div") %>% 
  as.data.frame() %>% 
  mutate(subsite = recode(subsite, "terrace" = "TE", "plateau" = "PL", 
                          "campinarana" = "C1","campina" = "C2")) 

fun_adiv <- read_excel("FUN_UNITE10.xlsx", sheet = "meta_div") %>% 
  as.data.frame() %>% 
  mutate(subsite = recode(subsite, "terrace" = "TE", "plateau" = "PL", 
                          "campinarana" = "C1","campina" = "C2")) 

## Group all possible combinations 
bac_adiv$Group <- with(bac_adiv, interaction(subsite, season, sep = "_"))
fun_adiv$Group <- with(fun_adiv, interaction(subsite, season, sep = "_"))

## Stats as letter display
anova_bac.shan <- aov(div_shan ~ Group, data = bac_adiv) # ANOVA
anova_fun.shan <- aov(div_shan ~ Group, data = fun_adiv) # ANOVA

tukey_bac.shan <- TukeyHSD(anova_bac.shan) # Tukey HSD Posthoc
tukey_fun.shan <- TukeyHSD(anova_fun.shan) # Tukey HSD Posthoc

library(multcompView)
cld_bac.shan <- multcompLetters4(anova_bac.shan, tukey_bac.shan) # letter display
cld_fun.shan <- multcompLetters4(anova_fun.shan, tukey_fun.shan) # letter display

cld_bac.shan <- data.frame(Group = names(cld_bac.shan$Group$Letters), 
                           Letters = cld_bac.shan$Group$Letters)
cld_fun.shan <- data.frame(Group = names(cld_fun.shan$Group$Letters), 
                           Letters = cld_fun.shan$Group$Letters)

letters_bac.shan <- tidyr::separate(cld_bac.shan, Group, 
                                    into = c("subsite", "season"), sep = "_")
letters_fun.shan <- tidyr::separate(cld_fun.shan, Group, 
                                    into = c("subsite", "season"), sep = "_")

letters_bac.shan$site_season <- rownames(letters_bac.shan)
letters_fun.shan$site_season <- rownames(letters_fun.shan)

## Plot alpha div bar plot 
library(dplyr)
library(ggpubr)

bac_adiv %>%
  mutate(Group = factor(Group, levels = c("PL_dry", "PL_wet", 
                                          "TE_dry", "TE_wet",
                                          "C1_dry", "C1_wet", 
                                          "C2_dry", "C2_wet"))) %>%  
  ggbarplot("Group","div_shan", fill = "Group", color = "Group", 
            palette = c("lightgreen","seagreen3", "lightgreen","seagreen3",
                        "bisque2", "bisque4","bisque2", "bisque4"),       
            position = position_dodge(0.8), 
            add = "mean_sd", error.plot = "upper_errorbar", width = 0.7) +
  geom_text(data = distinct(letters_bac.shan, site_season, Letters), 
            aes(x = site_season, y = max(bac_adiv$div_shan) + 0.2, 
                label = Letters), position = position_dodge(width = 0.8)) + 
  labs(x=NULL, y="Shannon index") +
  theme_pubr() + theme(panel.grid = element_blank(), legend.position = "top") +
  ggpubr::theme_pubr() + coord_flip() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "bottom", legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  # Add median as square
  stat_summary(aes(group = interaction(subsite, season)), 
               fun = median, geom = "point", shape = 0, size = 3, fill = "black",
               position = position_dodge(0.8)) +
  scale_x_discrete(labels = c("PL_dry" = "PL-dry","PL_wet" = "PL-wet",
                              "TE_dry" = "TE-dry","TE_wet" = "TE-wet",
                              "C1_dry" = "C1-dry","C1_wet" = "C1-wet",
                              "C2_dry" = "C2-dry","C2_wet" = "C2-wet"))

fun_adiv %>%
  mutate(Group = factor(Group, levels = c("PL_dry", "PL_wet", "TE_dry", "TE_wet",
                                          "C1_dry", "C1_wet", "C2_dry", "C2_wet"))) %>%  
  ggbarplot("Group","div_shan", fill = "Group", color = "Group", 
            palette = c("lightgreen","seagreen3","lightgreen","seagreen3",
                        "bisque2", "bisque4","bisque2", "bisque4"), 
            position = position_dodge(0.8), 
            add = "mean_sd", error.plot = "upper_errorbar", width = 0.7) +
  geom_text(data = distinct(letters_bac.shan, site_season, Letters), 
            aes(x = site_season, y = max(fun_adiv$div_shan) + 0.2, 
                label = Letters), position = position_dodge(width = 0.8)) + 
  labs(x=NULL, y="Shannon index") +
  theme_pubr() + theme(panel.grid = element_blank(), legend.position = "top") +
  ggpubr::theme_pubr() + coord_flip() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "bottom", legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  # Add median as square
  stat_summary(aes(group = interaction(subsite, season)), 
               fun = median, geom = "point", shape = 0, size = 3, fill = "black",
               position = position_dodge(0.8)) +
  scale_x_discrete(labels = c("PL_dry" = "PL-dry","PL_wet" = "PL-wet",
                              "TE_dry" = "TE-dry","TE_wet" = "TE-wet",
                              "C1_dry" = "C1-dry","C1_wet" = "C1-wet",
                              "C2_dry" = "C2-dry","C2_wet" = "C2-wet"))
```

### 3 \| Beta diversity analsis

#### 3.1 \| Prerequisite steps

Initially, we filtered out the sample `DSTB` as stated and discussed in the manuscript (see 1 \| Pre-processing). We then extract our feature and sample data, and introduce a grouping variable for later plotting based on grouping (e.g. forest type, season). Lastly, we define a common ordination plot theme for consistency.

```{r rda1, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Load dependencies
library(dplyr)
library(tidyr)
library(phyloseq)
library(vegan)

## Extract & transform data for beta diversity ordination
otu.bac = ps.meco.bac %>% phyloseq::otu_table() %>% 
  t() %>% decostand(method = "hellinger") %>% as.data.frame()

otu.fun = ps.meco.fun %>% phyloseq::otu_table() %>% 
  t() %>% decostand(method = "hellinger") %>% as.data.frame()

data.bac = data.frame(phyloseq::sample_data(ps.meco.bac))
data.fun = data.frame(phyloseq::sample_data(ps.meco.fun))

## Make grouping variable
data.bac$Group <- paste(data.bac$site, data.bac$season, sep = "_")
data.bac$Group <- factor(data.bac$Group, levels = c(
  "TF_dry", "TF_wet", "WSE_dry", "WSE_wet"))

data.fun$Group <- paste(data.fun$site, data.bac$season, sep = "_")
data.fun$Group <- factor(data.fun$Group, levels = c(
  "TF_dry", "TF_wet", "WSE_dry", "WSE_wet"))

## PCOA / RDA theme 
theme_coord <- function (base_size = 12, base_family = "", border = FALSE, margin = TRUE, 
                         legend = c("bottom"), x.text.angle = 0) 
{
  half_line <- base_size/2
  if (!is.numeric(legend)) 
    legend <- match.arg(legend)
  if (x.text.angle > 5) 
    xhjust <- 1
  else xhjust <- NULL
  if (margin) 
    plot.margin <- margin(half_line, half_line, half_line, half_line)
  else plot.margin <- unit(c(0.5, 0.3, 0.3, 0.3), "mm")
  .theme <- 
    theme_bw(base_size = base_size, base_family = base_family) %+replace% 
    theme(panel.border = element_rect(fill = NA, colour = "black", size = 0.7),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(), 
          axis.line = element_blank(), 
          axis.text = element_text(color = "black"), 
          legend.key = element_blank(), 
          strip.background = element_rect(fill = "#F2F2F2", colour = "black", size = 0.7), 
          plot.margin = plot.margin, 
          legend.position = legend, complete = TRUE,
          plot.title = element_text(hjust = 0.5, face = "bold", size = 16, vjust = 2))
  if (x.text.angle != 0) 
    .theme <- .theme + theme(axis.text.x = element_text(angle = x.text.angle, 
                                                        hjust = xhjust))
  .theme
}
```

#### 3.2 \| Permutational analysis of variance (PERMANOVA) w/ `vegan`

While this is a separate table in the respective manuscript (supplementary Table S2), we include PERMANOVA in this script as it is a vital part of microbial community diversity analysis. Downstream, we use the PERMANOVA results to directly plot with our ordinations. So, to show how this is done, we include PERMANOVA as part of this code script. Since we use hellinger-transformed counts for the later RDA, we use the same distance type (=*euclidean*) for PERMANOVA.

```{r perm1, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Calc distance 
library(phyloseq)
euc_dist.ps.bac <- vegan::vegdist(otu.bac, method = "euclidean")
euc_dist.ps.fun <- vegan::vegdist(otu.fun, method = "euclidean")

## Check interaction between predictors
betadisper(euc_dist.ps.bac, group = interaction(data.bac$season, data.bac$site))
betadisper(euc_dist.ps.fun, group = interaction(data.fun$season, data.fun$site))
```

We then sequentially fit all variables and identify which have significant impacts on microbial community composition. For later RDA we only keep the significant-effect variables as constraints.

```{r perm2, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## PERMANOVA: sequentially-fitting variables
perm.bac = vegan::adonis2(euc_dist.ps.bac ~ season * site + subsite, 
                          data = data.bac, permutations = 999, by = "terms")
perm.fun = vegan::adonis2(euc_dist.ps.fun ~ season * site + subsite, 
                          data = data.fun, permutations = 999, by = "terms")

perm.bac
perm.fun

## PERMANOVA: globally (same formula)
perm.bac.global = vegan::adonis2(euc_dist.ps.bac ~ season * site + subsite, 
                                 data = data.bac, permutations = 999)
perm.fun.global = vegan::adonis2(euc_dist.ps.fun ~ season * site + subsite, 
                                 data = data.fun, permutations = 999)
perm.bac.global
perm.fun.global

## Define PERMANOVA labels for ordination plots
perm.R2_bac <- round(perm.bac.global$R2[1], 3)
perm.R2_fun <- round(perm.fun.global$R2[1], 3)

perm.p_bac  <- round(perm.bac.global$`Pr(>F)`[1], 3)
perm.p_fun  <- round(perm.fun.global$`Pr(>F)`[1], 3)

perm.label_bac <- paste0("PERMANOVA:\nR^2 = ", perm.R2_bac, "\np = ", perm.p_bac)
perm.label_fun <- paste0("PERMANOVA:\nRB2 = ", perm.R2_fun, "\np = ", perm.p_fun)
```

Lastly, we calculate the %Sum of Squares for later reporting in the manuscript. For this purpose, we defined a customary function to directly apply to our PERMANOVA results and retrieve %SS values.

```{r perm3, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Define function to calc %SS
get_permanova_stats <- function(adonis_obj) {
  aov_tab <- adonis_obj
  # Extract SS including residuals for %SS
  total_SS <- sum(aov_tab$SumOfSqs, na.rm = TRUE)
  # Remove residuals for table display
  aov_tab <- aov_tab[!rownames(aov_tab) %in% c("Residual", "Total"), ]
  Fval <- aov_tab$F
  pval <- aov_tab$`Pr(>F)`
  perc_SS <- round((aov_tab$SumOfSqs / total_SS) * 100, 3)
  data.frame(
    Term = rownames(aov_tab),
    F_value = round(Fval, 3),
    p_value = round(pval, 3),
    Percent_SS = perc_SS,
    row.names = NULL
  )
}

## Calc %SS
print(get_permanova_stats(perm.bac))
print(get_permanova_stats(perm.fun))
```

#### 3.3 \| Redundancy analysis (RDA) w/ `vegan`

After testing which variables (e.g. season, forest type/site, sampling location/subsite) have a significant impact on community composition using PERMANOVA, we use these variables to constrain the RDA ordination.

```{r rda2, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Computed mixed models based on PERMANOVA results
rda_model.bac = vegan::rda(otu.bac ~ season * site + subsite, data = data.bac)
rda_model.fun = vegan::rda(otu.fun ~ season * site + subsite, data = data.fun)

## Calculate RDA axis scores
eigenvalues.bac <- summary(rda_model.bac)$concont$importance["Eigenvalue", ]
eigenvalues.fun <- summary(rda_model.fun)$concont$importance["Eigenvalue", ]
variance.bac <- eigenvalues.bac / sum(eigenvalues.bac)
variance.fun <- eigenvalues.fun / sum(eigenvalues.fun)

variance.bac = variance.bac %>% as.list()
variance.fun = variance.fun %>% as.list()

## RDA axis scores
variance.bac
variance.fun
```

Now, we can compute ordination based on the RDA models.

**Important!** RDA itself is – basically – a constrained Principal Component Analysis, operating in a mathematically Euclidean space. Hence, it requires data to be normalized and transformed accordingly. If we want to use RDA, we thus have to apply either CLR transformation, rCLR transformation, or hellinger transformation, the latter being really common. Using rCLR transformation will evaluate robust Aitchison (rAitchison) distance instead. The alternative to PCA/RDA that operate in an Euclidean space is Principal Coordinate Analysis (PCoA), which is based on e.g. Bray-Curtis dissimilarity, and is commonly used in ecological studies.

```{r rda3, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Extract ordination scores
scores.bac <- scores(rda_model.bac, display = "sites")
scores.fun <- scores(rda_model.fun, display = "sites")

## Compute RDA ordination
ordination.rda_bac <- data.frame(scores.bac, 
                                 season = data.bac$season, 
                                 site = data.bac$site, 
                                 subsite = data.bac$subsite)
ordination.rda_fun <- data.frame(scores.fun, 
                                 season = data.fun$season, 
                                 site = data.fun$site, 
                                 subsite = data.fun$subsite)

ordination.rda_bac$Group <- data.bac$Group
ordination.rda_fun$Group <- data.fun$Group
ordination.rda_bac$smpID <- rownames(data.bac)
ordination.rda_fun$smpID <- rownames(data.fun)

## Plot RDA 
library(ggplot2) 

ggplot(ordination.rda_bac, aes(x=RDA1, y=RDA2, color = Group, shape = subsite)) + 
  geom_point(size = 5, alpha = 0.5) + theme_coord() + 
  scale_color_manual(values = c("TF_dry" = "lightgreen", "TF_wet" = "forestgreen", 
                                "WSE_dry" = "bisque", "WSE_wet" = "bisque4")) +
  scale_fill_manual(values = c("TF_dry" = "lightgreen", "TF_wet" = "forestgreen", 
                               "WSE_dry" = "bisque", "WSE_wet" = "bisque4")) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") + 
  scale_shape_manual(values=c(18, 15, 17, 16)) +
  labs(title = "16S-ASVs", 
       x = paste0("RDA1 (", round(variance.bac$RDA1*100, 2), "%)"), 
       y = paste0("RDA2 (", round(variance.bac$RDA2*100, 2), "%)")) +
  annotate("text", 
           x = min(ordination.rda_bac$RDA1) * 1, 
           y = min(ordination.rda_bac$RDA2) * 0.8, 
           label = perm.label_bac, hjust = 0, size = 4) +
  scale_x_continuous(limits = c(-0.85, 0.55), 
                     labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  scale_y_continuous(limits = c(-0.55, 0.7), 
                     labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  geom_text(aes(label = smpID), check_overlap = TRUE) 
  
ggplot(ordination.rda_fun, aes(x=RDA1, y=RDA2, color = Group, shape = subsite)) + 
  geom_point(size = 5, alpha = 0.5) + theme_coord() + 
  scale_color_manual(values = c("TF_dry" = "lightgreen", "TF_wet" = "forestgreen", 
                                "WSE_dry" = "bisque", "WSE_wet" = "bisque4")) +
  scale_fill_manual(values = c("TF_dry" = "lightgreen", "TF_wet" = "forestgreen", 
                               "WSE_dry" = "bisque", "WSE_wet" = "bisque4")) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") + 
  scale_shape_manual(values=c(18, 15, 17, 16)) +
  labs(title = "ITS-ASVs", 
       x = paste0("RDA1 (", round(variance.fun$RDA1*100, 2), "%)"), 
       y = paste0("RDA2 (", round(variance.fun$RDA2*100, 2), "%)")) +
  annotate("text", 
           x = min(ordination.rda_fun$RDA1) * 1, 
           y = min(ordination.rda_fun$RDA2) * 0.8, 
           label = perm.label_fun, hjust = 0, size = 4) +
  scale_x_continuous(limits = c(-0.52,0.7), 
                     labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  scale_y_continuous(limits = c(-0.85,0.5), 
                     labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  geom_text(aes(label = smpID), check_overlap = TRUE) 
```

#### 3.4 \| Principal coordinate analysis (PCoA) w/ `vegan`

While we used RDA for our manuscript, PCoA can also be used, and is commonly used, to visualize beta diversity of microbial communities. Hence, we provide a comprehensive script to run PCoA too. Here, we chose Bray-Curtis dissimilarity matrix, which is calculated based on normalized (total sum scaling) counts.

```{r pcoa, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(phyloseq)
library(vegan)
library(ggplot2)
library(ggrepel)

## Calculate distance matrix
dist_bac <- phyloseq::distance(ps.bac.tss, method = "bray")
dist_fun <- phyloseq::distance(ps.fun.tss, method = "bray")

## Calculate ordination based on dist
pcoa_bac <- ordinate(ps.bac.tss, method = "PCoA", distance = dist_bac)
pcoa_fun <- ordinate(ps.fun.tss, method = "PCoA", distance = dist_fun)

## Plot PCoA
plot_ordination(ps.bac.tss, pcoa_bac, color = "Group", shape = "subsite") +
  geom_point(aes(size = 5, alpha = 0.5)) + 
  theme_coord() + 
  labs(title = "16S-ASVs") + 
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") +
  scale_shape_manual(values = c(18, 15, 17, 16)) +
  annotate("text", 
           x = min(pcoa_bac$Axis.1) * 0.75, 
           y = min(pcoa_bac$Axis.2) * 0.325, 
           label = perm.label_bac, hjust = 0, size = 4)

plot_ordination(ps.fun.tss, pcoa_fun, color = "Group", shape = "subsite") +
  geom_point(aes(size = 5, alpha = 0.5)) + 
  theme_coord() + 
  labs(title = "ITS-ASVs") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") +
  scale_shape_manual(values = c(18, 15, 17, 16)) +
  annotate("text", 
           x = min(pcoa_fun$Axis.1) * 0.6, 
           y = min(pcoa_fun$Axis.2) * 0.4, 
           label = perm.label_fun, hjust = 0, size = 4) 
```

#### 3.5 \| Beta diversity distance

After importing our data, we'll apply the filter to analyse only putative SOM-cycling taxa, creating a joint bacteria-fungi df for subsequent plotting with `ggplot2`.

```{r betadiv1, message=FALSE, warning=FALSE, results='hide', tidy=FALSE}
library(vegan)       # for distance matrix calculation
library(ggplot2)     # for plotting
library(dplyr)       # for data wrangling
library(tidyr)       # for data wrangling
library(ggsignif)    # optional, for significance marks

## Already exists from step (1 | Pre-processing) 
otu.bac # hellinger-transformed counts
otu.fun # hellinger-transformed counts 
dist_bac.hell <- vegdist(otu.bac, method = "euclidean")
dist_fun.hell <- vegdist(otu.fun, method = "euclidean")

## Create grouping variables
data.bac$Group <- paste(data.bac$site, data.bac$season, sep = "_")
data.bac$Group <- factor(data.bac$Group, levels = c("TF_dry", "TF_wet", "WSE_dry", "WSE_wet"))

data.fun$Group <- paste(data.fun$site, data.bac$season, sep = "_")
data.fun$Group <- factor(data.fun$Group, levels = c("TF_dry", "TF_wet", "WSE_dry", "WSE_wet"))

data.bac$Group2 <- paste(data.bac$subsite, data.bac$season, sep = "_")
data.bac$Group2 <- factor(data.bac$Group2, levels = c(
  "plateau_dry","campinarana_dry","campina_dry","terrace_dry",
  "campina_wet","campinarana_wet","plateau_wet","terrace_wet"))

data.fun$Group2 <- paste(data.fun$subsite, data.bac$season, sep = "_")
data.fun$Group2 <- factor(data.fun$Group2, levels = c(
  "plateau_dry","campinarana_dry","campina_dry","terrace_dry",
  "campina_wet","campinarana_wet","plateau_wet","terrace_wet"))

## Create dist_df long-format 
dist_bac.df <- as.data.frame(as.table(as.matrix(dist_bac.hell))) %>%
  rename(sample1 = Var1, sample2 = Var2, distance = Freq) %>%
  filter(as.character(sample1) < as.character(sample2)) %>%  # upper triangle only
  mutate(group1 = data.bac[sample1, "Group"], group2 = data.bac[sample2, "Group"]) %>%
  # keep only within-group distances
  filter(group1 == group2) %>%
  select(distance, group = group1)

dist_fun.df <- as.data.frame(as.table(as.matrix(dist_fun.hell))) %>%
  rename(sample1 = Var1, sample2 = Var2, distance = Freq) %>%
  filter(as.character(sample1) < as.character(sample2)) %>%
  mutate(group1 = data.fun[sample1, "Group"], group2 = data.fun[sample2, "Group"]) %>%
  filter(group1 == group2) %>%
  select(distance, group = group1)

dist_bac.df2 <- as.data.frame(as.table(as.matrix(dist_bac.hell))) %>%
  rename(sample1 = Var1, sample2 = Var2, distance = Freq) %>%
  filter(as.character(sample1) < as.character(sample2)) %>%
  mutate(group1 = data.bac[sample1, "Group2"], group2 = data.bac[sample2, "Group2"]) %>%
  filter(group1 == group2) %>%
  select(distance, group = group1)

dist_fun.df2 <- as.data.frame(as.table(as.matrix(dist_fun.hell))) %>%
  rename(sample1 = Var1, sample2 = Var2, distance = Freq) %>%
  filter(as.character(sample1) < as.character(sample2)) %>%
  mutate(group1 = data.fun[sample1, "Group2"], group2 = data.fun[sample2, "Group2"]) %>%
  filter(group1 == group2) %>%
  select(distance, group = group1)

## Calculate stats letters
library(multcompView) 
anova_bac <- aov(distance ~ group, data = dist_bac.df)
tukey_bac <- TukeyHSD(anova_bac)
letters_bac <- multcompLetters4(anova_bac, tukey_bac)
anova_fun <- aov(distance ~ group, data = dist_fun.df)
tukey_fun <- TukeyHSD(anova_fun)
letters_fun <- multcompLetters4(anova_fun, tukey_fun)
letters_bac$group
letters_fun$group

## Set plotting order & rename to desired labels 
dist_bac.df$group <- factor(dist_bac.df$group, 
                            levels = c("TF_dry","TF_wet","WSE_dry","WSE_wet"),
                            labels = c("TF-dry","TF-wet","WS-dry","WS-wet"))
dist_fun.df$group <- factor(dist_fun.df$group, 
                            levels = c("TF_dry","TF_wet","WSE_dry","WSE_wet"),
                            labels = c("TF-dry","TF-wet","WS-dry","WS-wet"))

dist_bac.df2$group <- factor(dist_bac.df2$group, 
                             levels = c("plateau_dry","plateau_wet",
                                        "terrace_dry","terrace_wet",
                                        "campinarana_dry","campinarana_wet",
                                        "campina_dry","campina_wet"),
                             labels = c("PL-dry","PL-wet","TE-dry","TE-wet",
                                        "C1-dry","C1-wet","C2-dry","C2-wet"))
dist_fun.df2$group <- factor(dist_fun.df2$group, 
                             levels = c("plateau_dry","plateau_wet",
                                        "terrace_dry","terrace_wet",
                                        "campinarana_dry","campinarana_wet",
                                        "campina_dry","campina_wet"),
                             labels = c("PL-dry","PL-wet","TE-dry","TE-wet",
                                        "C1-dry","C1-wet","C2-dry","C2-wet"))

## Plot as bxp (resolution = forest type)
library(ggplot2)
library(scales)

dist_bac.df %>%
  ggplot(aes(x = group, y = distance, fill = group)) + 
  geom_boxplot() + 
  ggpubr::theme_pubr() + 
  scale_fill_manual(values = c(
    "TF-dry" = "lightgreen","WS-dry"= "bisque",
    "WS-wet"= "bisque4","TF-wet"= "forestgreen")) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  ggpubr::theme_pubr() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 0), 
        legend.position = "bottom", legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  labs(x=NULL, y="Hellinger distance") +
  geom_jitter(aes(shape = subsite), width = 0.2, size = 2.5, color = "black", alpha = 0.5) +
    scale_shape_manual(values = c("PL" = 17,"TE" = 16,"C1" = 15,"C2" = 18)) +
  dist_bac.df2 %>%
  rename(group2 = group) %>%
  mutate(group = case_when(group2 %in% c("PL-wet", "TE-wet") ~ "TF-wet", 
                           group2 %in% c("PL-dry", "TE-dry") ~ "TF-dry",
                           group2 %in% c("C1-dry", "C2-dry") ~ "WS-dry", 
                           group2 %in% c("C1-wet", "C2-wet") ~ "WS-wet", 
                           TRUE ~ as.character(group2)),
         subsite = case_when(grepl("^PL", group2) ~ "PL",
                             grepl("^TE", group2) ~ "TE",
                             grepl("^C1", group2) ~ "C1",
                             grepl("^C2", group2) ~ "C2",TRUE ~ NA_character_)) 

dist_fun.df %>%
  ggplot(aes(x = group, y = distance, fill = group)) + 
  geom_boxplot() + 
  ggpubr::theme_pubr() + 
  scale_fill_manual(values = c(
    "TF-dry" = "lightgreen","WS-dry"= "bisque",
    "WS-wet"= "bisque4","TF-wet"= "forestgreen")) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  # Panel border black
  ggpubr::theme_pubr() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 0), 
        legend.position = "bottom", 
        legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  labs(x=NULL, y="Hellinger distance") +
  geom_jitter(aes(shape = subsite), width = 0.2, size = 2.5, color = "black", alpha = 0.5) +
  scale_shape_manual(values = c("PL" = 17,"TE" = 16,"C1" = 15,"C2" = 18)) +
  dist_fun.df2 %>%
  rename(group2 = group) %>%
  mutate(group = case_when(group2 %in% c("PL-wet", "TE-wet") ~ "TF-wet", 
                           group2 %in% c("PL-dry", "TE-dry") ~ "TF-dry",
                           group2 %in% c("C1-dry", "C2-dry") ~ "WS-dry", 
                           group2 %in% c("C1-wet", "C2-wet") ~ "WS-wet", 
                           TRUE ~ as.character(group2)),
         subsite = case_when(grepl("^PL", group2) ~ "PL",
                             grepl("^TE", group2) ~ "TE",
                             grepl("^C1", group2) ~ "C1",
                             grepl("^C2", group2) ~ "C2",TRUE ~ NA_character_))
```

### 4 \| Other analyses

#### 4.1 \| Beta nearest taxon index (beta-NTI) w/ `picante`

**What is the beta-NTI?** It's a z-score representing how far the observed β-MNTD deviates from the null model expectation, with a \|ß-NTI\| \> \|2\| indicating that communities are more phylogenetically different than expected (heterogeneous selection) as a result of primarily deterministic assembly processes, while a \|ß-NTI\| \< \|2\| indiates that communities are more similar than expected (homogeneous selection), due to stochastic assembly processes.

In our manuscript, we used beta-NTI to compare forest groups, and whether community assembly was driven deterministically or stochastically within seasons. Hence, we first need to prepare our data again by introducing the respective grouping parameters.

```{r bnti1, message=FALSE, warning=FALSE, results='hide', tidy=FALSE}
## Load dependencies
library(phyloseq)
library(vegan)
library(tidyverse)

ps.meco.bac # see 1|Pre-processing
ps.meco.fun # see 1|Pre-processing

## Define forest type & season as labels 
forest_col <- "site"
season_col <- "season"

tf_label  <- "TF"
wse_label <- "WSE"

## Subset phyloseq object by forest type 
## - this subsetting is derived from metadata 
## - leading to potential empty/0 taxa abundances
## - these have to be subsequently removed 

## ## Bacteria 
ps.meco.bac_tf <- subset_samples(ps.meco.bac, get_variable(ps.meco.bac, forest_col) == tf_label)
ps.meco.bac_tf <- prune_taxa(taxa_sums(ps.meco.bac_tf)  > 0, ps.meco.bac_tf)

ps.meco.bac_ws <- subset_samples(ps.meco.bac, get_variable(ps.meco.bac, forest_col) == wse_label)
ps.meco.bac_ws <- prune_taxa(taxa_sums(ps.meco.bac_ws) > 0, ps.meco.bac_ws)

ps.meco.fun_tf <- subset_samples(ps.meco.fun, get_variable(ps.meco.fun, forest_col) == tf_label)
ps.meco.fun_tf <- prune_taxa(taxa_sums(ps.meco.fun_tf)  > 0, ps.meco.fun_tf)

ps.meco.fun_ws <- subset_samples(ps.meco.fun, get_variable(ps.meco.fun, forest_col) == wse_label)
ps.meco.fun_ws <- prune_taxa(taxa_sums(ps.meco.fun_ws) > 0, ps.meco.fun_ws)
```

Now, we can compute the beta-NTI using the `picante` package. This is done by random tree tip shuffling, aka the phyloseq object requires a `phylo tree` – phylogenetic tree information. To simplify beta-NTI calculations, we defined a customary automated function, applicable to our subsetted `phyloseq` objects to calculate and visualize beta-NTI. The steps in the function are:

1.  EXTRACT NECESSARY MATRICES

    -   This is necessary for `picante::comdistnt` however, `picante::comdistnt` expects ASVs as columns
    -   As our data currently has ASVs as rows, we need to transpose the extracted ASV table
    -   Additionally, we must extract the phylogenetic tree, which is needed to calculate the beta-NTI
    -   And we need to extract the sample_data because this has the grouping variable information (e.g. season)

2.  SANITY CHECK

    -   This step ensures we can run the function smoothly by checking if the `phylo tree` matches the ASVs
    -   It makes sure the tree exists and is a proper `phylo tree`

3.  MATCH ASVs TO TREE TIPS *(vital to β-NTI calculation)*

    -   Matches ASV table-tree tip intersection to ensure that only taxa present in both are used for calculations
    -   Identified taxa are defined as "common" (or shared) taxa
    -   To avoid later problems we further subset our `phyloseq` object to only shared taxa

4.  COMPUTE NTI VARIABLES

    -   This step first calculates the *β-MNTD* (Mean Nearest Taxon Distance between communities)
    -   This first part measures how phylogenetically different two samples are on average weighted by their relative abundances
    -   Based on these calculations, a null model with repeated randomizations at the tree tips is created to investgígate how the *β-MNTD* changes under random phylogenetic relationships
    -   This breaks the phylogenetic structure while keeping abundance structures
    -   Using the null model, it then calulcates the mean and standard deviation of the null distances
    -   Lastly, the *β-NTI* is calculated by substracting the *β-MNTD* with the mean of the distances, and dividing it with the standard deviation

**Important!** We calculate beta-NTI between seasons for each forest type separately, and later combine the two forest types into one plot as shown in the manuscript.

```{r bnti2, message=FALSE, warning=FALSE, results='hide', tidy=TRUE}
## Load dependencies
library(picante) # for ses.mntd and beta.mntd
library(ape)

## Write function to compute βNTI
compute_betaNTI.season <- function(ps, nreps = 999) {
  # STEP 1: Extract OTU table
  otu <- as(otu_table(ps), "matrix")
  if (taxa_are_rows(ps)) otu <- t(otu)
  otu <- otu[rowSums(otu) > 0, , drop = FALSE]
  samples <- rownames(otu)
  
  # STEP 2: Extract tree & match taxa
  tree <- phy_tree(ps)
  if (is.null(tree)) stop("No phylogenetic tree found in phyloseq object.")
  shared <- intersect(colnames(otu), tree$tip.label)
  if (length(shared) < 3) stop("Too few overlapping taxa between OTU table and tree.")
  
  otu <- otu[, shared, drop = FALSE]
  tree <- ape::keep.tip(tree, shared)
  otu <- otu[, tree$tip.label, drop = FALSE]
  
  # STEP 3: Observed βMNTD
  beta_mntd_obs <- picante::comdistnt(otu, cophenetic(tree), abundance.weighted = TRUE)
  beta_mntd_obs <- as.matrix(beta_mntd_obs)
  samples <- rownames(beta_mntd_obs)
  beta_mntd_obs <- beta_mntd_obs[samples, samples]
  
  # STEP 4: Null model (manual repetitions)
  null_list <- list()
  for (i in seq_len(nreps)) {
    tree_rand <- picante::tipShuffle(tree)
    mat <- picante::comdistnt(otu, cophenetic(tree_rand), abundance.weighted = TRUE)
    mat <- as.matrix(mat)
    mat <- mat[rownames(beta_mntd_obs), colnames(beta_mntd_obs)]
    null_list[[i]] <- mat
  }
  
  # STEP 5: Compute mean and SD
  mean_null <- Reduce("+", null_list) / nreps
  sd_null <- sqrt(Reduce("+", lapply(null_list, function(x) (x - mean_null)^2)) / nreps)
  
  # STEP 6: Compute βNTI
  betaNTI <- (beta_mntd_obs - mean_null) / sd_null
  
  # STEP 7: Subset by season
  sd_meta <- as(sample_data(ps), "data.frame")
  if (!(season_col %in% colnames(sd_meta))) stop(paste0("Column '", season_col, "' not found in sample_data."))
  
  drought_samples <- rownames(sd_meta)[sd_meta[[season_col]] == "dry"]
  control_samples <- rownames(sd_meta)[sd_meta[[season_col]] == "wet"]
  
  # STEP 8: Prepare output table
  library(reshape2)
  betaNTI_df <- melt(betaNTI, varnames = c("Sample1", "Sample2"), value.name = "betaNTI")
  
  betaNTI_df$Season1 <- sd_meta[[season_col]][match(betaNTI_df$Sample1, rownames(sd_meta))]
  betaNTI_df$Season2 <- sd_meta[[season_col]][match(betaNTI_df$Sample2, rownames(sd_meta))]
  
  betaNTI_df$Comparison <- NA
  betaNTI_df$Comparison[betaNTI_df$Season1 == "dry" & betaNTI_df$Season2 == "dry"] <- "Dry (Within-season)"
  betaNTI_df$Comparison[betaNTI_df$Season1 == "wet" & betaNTI_df$Season2 == "wet"] <- "Wet (Within-season)"
  betaNTI_df$Comparison[betaNTI_df$Season1 != betaNTI_df$Season2] <- "Between-season"
  
  # STEP 9: Fraction deterministic
  drought_vs_control <- betaNTI[drought_samples, control_samples]
  frac_deterministic <- sum(abs(drought_vs_control) > 2, na.rm = TRUE) / length(drought_vs_control)
  message(sprintf("Fraction deterministic (|βNTI| > 2) = %.2f", frac_deterministic))
  
  # Return tidy βNTI table
  return(betaNTI_df)
}

## Calculate βNTI within each forest type 
betaNTI.bac_tf  = compute_betaNTI.season(ps.meco.bac_tf, nreps = 999)
betaNTI.bac_wse = compute_betaNTI.season(ps.meco.bac_ws, nreps = 999)

betaNTI.fun_tf  = compute_betaNTI.season(ps.meco.fun_tf, nreps = 999)
betaNTI.fun_wse = compute_betaNTI.season(ps.meco.fun_ws, nreps = 999)
```

After computing the beta-NTI of each season per forest type, we have to clean up the beta-NTI dataframe for subsequent plotting. We use another customary function to do so. The individual steps are found in the code below.

```{r bnti3, message=FALSE, warning=FALSE, results='hide', tidy=TRUE}
## Write function to clean up βNTI dataframe 
clean_betaNTI.season <- function(ps, betaNTI_df) {
  # Ensure sample_data is a data frame
  sd <- as(sample_data(ps), "data.frame")
  if (!(season_col %in% colnames(sd))) {
    stop(paste0("Column '", season_col, "' not found in sample_data."))
  }
  
  # Define sample groups
  drought_samples <- rownames(sd)[sd[[season_col]] == "dry"]
  control_samples <- rownames(sd)[sd[[season_col]] == "wet"]
  all_samples <- c(drought_samples, control_samples)
  
  # Filter to only relevant samples
  betaNTI_df <- subset(betaNTI_df, Sample1 %in% all_samples & Sample2 %in% all_samples)
  
  # Remove self-comparisons (Sample1 == Sample2)
  betaNTI_df <- betaNTI_df[betaNTI_df$Sample1 != betaNTI_df$Sample2, ]
  
  # Remove duplicate pairs (since βNTI is symmetric)
  betaNTI_df <- betaNTI_df[!duplicated(t(apply(betaNTI_df[, 1:2], 1, sort))), ]
  
  # Return cleaned data frame
  return(betaNTI_df)
}

## Clean up βNTI dataframe for plotting 
betaNTI.clean.bac_tf = clean_betaNTI.season(ps.meco.bac_tf, betaNTI.bac_tf)
betaNTI.clean.bac_wse = clean_betaNTI.season(ps.meco.bac_ws, betaNTI.bac_wse)
betaNTI.clean.fun_tf = clean_betaNTI.season(ps.meco.fun_tf, betaNTI.fun_tf)
betaNTI.clean.fun_wse = clean_betaNTI.season(ps.meco.fun_ws, betaNTI.fun_wse)
```

Lastly, we use another customary function to prepare the data for plotting, and then plot our beta-NTIs w/ `ggplot2`.

```{r bnti4, message=FALSE, warning=FALSE, results='hide', tidy=TRUE}
## Prepare data for plotting 
plot_betaNTI.by.season <- function(ps, betaNTI_df, group_col = "season") {
  library(ggplot2)
  sd <- as(sample_data(ps), "data.frame")
  
  # Ensure group column exists
  if (!(group_col %in% colnames(sd))) {
    stop(paste0("Column '", group_col, "' not found in sample_data."))
  }
  
  # Assign group info to each sample pair
  betaNTI_df$Group1 <- sd[[group_col]][match(betaNTI_df$Sample1, rownames(sd))]
  betaNTI_df$Group2 <- sd[[group_col]][match(betaNTI_df$Sample2, rownames(sd))]
  
  # Keep only within-group comparisons (both samples in the same group)
  betaNTI_within <- subset(betaNTI_df, Group1 == Group2)
  
  # Assign group label
  betaNTI_within$Group <- betaNTI_within$Group1
  
  return(betaNTI_within)
}

betaNTI.clean.plot_bac.tf = plot_betaNTI.by.season(ps.meco.bac_tf, betaNTI.clean.bac_tf, group_col = "season")
betaNTI.clean.plot_bac.wse = plot_betaNTI.by.season(ps.meco.bac_ws, betaNTI.clean.bac_wse, group_col = "season")
betaNTI.clean.plot_fun.tf = plot_betaNTI.by.season(ps.meco.fun_tf, betaNTI.clean.fun_tf, group_col = "season")
betaNTI.clean.plot_fun.wse = plot_betaNTI.by.season(ps.meco.fun_ws, betaNTI.clean.fun_wse, group_col = "season")

## Plot boxplot
library(ggplot2)
 
ggplot(betaNTI.clean.plot_bac.tf, aes(x=Group, y=betaNTI)) +
  geom_boxplot(outlier.colour = "red") + 
  geom_hline(yintercept=c(-2,2), linetype="dashed", color="red") +
  ylab("βNTI") + xlab("") + 
  ggpubr::theme_pubr() + 
  coord_cartesian(ylim = c(-5,3)) +
  geom_jitter(color="black", alpha = 0.9, width = 0.1)

ggplot(betaNTI.clean.plot_bac.wse, aes(x=Group, y=betaNTI)) +
  geom_boxplot(outlier.colour = "red") + 
  geom_hline(yintercept=c(-2,2), linetype="dashed", color="red") +
  ylab("βNTI") + xlab("") + 
  ggpubr::theme_pubr() + 
  coord_cartesian(ylim = c(-5,3)) +
  geom_jitter(color="black", alpha = 0.9, width = 0.1)

ggplot(betaNTI.clean.plot_fun.tf, aes(x=Group, y=betaNTI)) +
  geom_boxplot(outlier.colour = "red") + 
  geom_hline(yintercept=c(-2,2), linetype="dashed", color="red") +
  ylab("βNTI") + xlab("") + 
  ggpubr::theme_pubr() + 
  coord_cartesian(ylim = c(-5,3)) +
  geom_jitter(color="black", alpha = 0.9, width = 0.1)

ggplot(betaNTI.clean.plot_fun.wse, aes(x=Group, y=betaNTI)) +
  geom_boxplot(outlier.colour = "red") + 
  geom_hline(yintercept=c(-2,2), linetype="dashed", color="red") +
  ylab("βNTI") + xlab("") + 
  ggpubr::theme_pubr() + coord_cartesian(ylim = c(-5,3)) +
  geom_jitter(color="black", alpha = 0.9, width = 0.1)
```

#### 4.2 \| Beta diversity dispersion w/ `vegan`

Beta diversity dispersion – aka the homogeneity of multivariate dispersions – measures *how much species composition varies within a group*. Beta dispersion is calculated as the average distance from individual samples to their group centroid or median. It is a crucial tool in ecological and microbiome studies for comparing beta diversity variability between groups, but is not to be confused with distance variability in the data.

**Important!** We did not include this in the manuscript, but found it a useful tool to introduce under the umbrella of microbial diversity/community analysis regardless.

```{r betadisper1, message=FALSE, warning=FALSE, results='hide', tidy=FALSE}
## Define grouping column
data.bac2 = data.bac %>%
  mutate(site = recode(site, "WSE" = "WS")) %>%
  mutate(Group = paste(site, season, sep = "_"))
data.bac2$Group <- as.factor(data.bac2$Group)

data.fun2 = data.fun %>%
  mutate(site = recode(site, "WSE" = "WS")) %>%
  mutate(Group = paste(site, season, sep = "_"))
data.fun2$Group <- as.factor(data.fun2$Group)

## Calculate beta dispersion
## - the distances these are based on are the same as used for PERMANOVA (see 3.2)
## - in this case: hellinger-based euclidean distances 
library(vegan)
bd_bac <- betadisper(euc_dist.ps.bac, group = data.bac2$Group)
bd_fun <- betadisper(euc_dist.ps.fun, group = data.fun2$Group)

## Extract distance to centroid 
bd_dist.bac <- data.frame(distance = bd_bac$distances, group = bd_bac$group)
bd_dist.fun <- data.frame(distance = bd_fun$distances, group = bd_fun$group)

## Run stats w/ ANOVA-TukeyHSD
anova_bd.bac <- aov(distance ~ group, data = bd_dist.bac)
anova_bd.fun <- aov(distance ~ group, data = bd_dist.fun)

perm_bd.bac  <- permutest(bd_bac, permutations = 999)
perm_bd.fun  <- permutest(bd_fun, permutations = 999)

tukey_bd.bac <- TukeyHSD(anova_bd.bac)
tukey_bd.fun <- TukeyHSD(anova_bd.fun)

## Extract stats as letter display
library(multcompView)
letters_bd.bac <- multcompLetters4(anova_bd.bac, tukey_bd.bac)
letters_bd.fun <- multcompLetters4(anova_bd.fun, tukey_bd.fun)

## Set plotting order
bd_dist.bac$group <- factor(bd_dist.bac$group, 
                            levels = c("TF_dry","TF_wet","WS_dry","WS_wet"),
                            labels = c("TF-dry","TF-wet","WS-dry","WS-wet"))
bd_dist.fun$group <- factor(bd_dist.fun$group, 
                            levels = c("TF_dry","TF_wet","WS_dry","WS_wet"),
                            labels = c("TF-dry","TF-wet","WS-dry","WS-wet"))

## Plot beta dispersion per group
library(dplyr)
library(ggplot2)
library(ggpubr)

bd_dist.bac %>%
  ggplot(aes(x = group, y = distance, fill = group)) + 
  geom_boxplot() + ggpubr::theme_pubr() + labs(x=NULL, y="Distance") +
  scale_fill_manual(values = c("TF-dry" = "lightgreen","WS-dry"= "bisque",
                               "WS-wet"= "bisque4","TF-wet"= "forestgreen")) +
    scale_y_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  ggpubr::theme_pubr() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "bottom", 
        legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  labs(y="Beta diversity dispersion")

bd_dist.fun %>%
  ggplot(aes(x = group, y = distance, fill = group)) + 
  geom_boxplot() + ggpubr::theme_pubr() + labs(x=NULL, y="Distance") +
  scale_fill_manual(values = c("TF-dry" = "lightgreen","WS-dry"= "bisque",
                               "WS-wet"= "bisque4","TF-wet"= "forestgreen")) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  ggpubr::theme_pubr() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.75),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "bottom", 
        legend.title = element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face = "bold", size = 12)) +
  labs(y="Beta diversity dispersion")
```
