---
title: "Fig. S3: Relative abundance of dominant bacterial and fungal taxa across two rainforest ecosystems."
author: "Jessica Finck"
date: "2026-02-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

### 1 \| Pre-processing: Define `phyloseq` objects for bacteria and fungi

Prior to analysis, we create a `phyloseq` object from our data. This will be used for most of the downstream analysis.

**Important!** The file 'rooted_tree.nwk' is generated via QIIME2 pre-processing workflow. The detailed script can be [**found here**](https://github.com/JFinck1994/general_scripts/blob/main/qiime2_amplicon-16S-PE-read-processing.sh)**.**

```{r phyloseq, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Load dependencies
library(readxl) # import xlsx files
library(ape)    # import nwk files
library(dplyr)  # data wrangling

## Import data 
otu_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "asv") %>% as.data.frame()
tax_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "tax") %>% as.data.frame()
data_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_bac <- ape::read.tree("BAC_rooted_tree_dada2.nwk")

otu_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "asv") %>% as.data.frame()
tax_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "tax") %>% as.data.frame()
data_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_fun <- ape::read.tree("FUN_rooted_tree_dada2.nwk")

## Format variable classes
library(dplyr)  # data wrangling

data_bac$site <- as.factor(data_bac$site)
data_bac$subsite <- as.factor(data_bac$subsite)
data_bac$season <- as.factor(data_bac$season)

data_fun$site <- as.factor(data_fun$site)
data_fun$subsite <- as.factor(data_fun$subsite)
data_fun$season <- as.factor(data_fun$season)

## Adjust taxa names  
tax_bac$Kingdom = gsub("d__","",as.character(tax_bac$Kingdom))
tax_bac$Phylum = gsub("p__","",as.character(tax_bac$Phylum))
tax_bac$Class = gsub("c__","",as.character(tax_bac$Class))
tax_bac$Order = gsub("o__","",as.character(tax_bac$Order))
tax_bac$Family = gsub("f__","",as.character(tax_bac$Family))
tax_bac$Genus = gsub("g__","",as.character(tax_bac$Genus))
tax_bac$Species = gsub("s__","",as.character(tax_bac$Species))
otu_bac <- otu_bac %>% tibble::column_to_rownames("asv")
tax_bac <- tax_bac %>% tibble::column_to_rownames("asv")
data_bac <- data_bac %>% tibble::column_to_rownames("smp_id")

tax_fun$Kingdom = gsub("k__","",as.character(tax_fun$Kingdom))
tax_fun$Phylum = gsub("p__","",as.character(tax_fun$Phylum))
tax_fun$Class = gsub("c__","",as.character(tax_fun$Class))
tax_fun$Order = gsub("o__","",as.character(tax_fun$Order))
tax_fun$Family = gsub("f__","",as.character(tax_fun$Family))
tax_fun$Genus = gsub("g__","",as.character(tax_fun$Genus))
tax_fun$Species = gsub("s__","",as.character(tax_fun$Species))

otu_fun <- otu_fun %>% tibble::column_to_rownames("asv")
tax_fun <- tax_fun %>% tibble::column_to_rownames("asv")
data_fun <- data_fun %>% tibble::column_to_rownames("smp_id")

## Assemble ps objects 
library(phyloseq)

otu_bac <- as.matrix(otu_bac)
tax_bac <- as.matrix(tax_bac)
OTU_B = phyloseq::otu_table(otu_bac, taxa_are_rows = T)
TAX_B = phyloseq::tax_table(tax_bac)
samples_B = phyloseq::sample_data(data_bac)
tree_B = phyloseq::phy_tree(tree_bac)
ps.bac <- phyloseq::phyloseq(OTU_B, TAX_B, samples_B, tree_B)

otu_fun <- as.matrix(otu_fun)
tax_fun <- as.matrix(tax_fun)
OTU_F = phyloseq::otu_table(otu_fun, taxa_are_rows = T)
TAX_F = phyloseq::tax_table(tax_fun)
samples_F = phyloseq::sample_data(data_fun)
tree_F = phyloseq::phy_tree(tree_fun)
ps.fun <- phyloseq::phyloseq(OTU_F, TAX_F, samples_F, tree_F)

ps.bac
ps.fun

## Tidy ps objects w/ microeco
library(microeco)
library(file2meco)

meco.fun = phyloseq2meco(ps.fun)
meco.bac = phyloseq2meco(ps.bac)

meco.bac$tax_table %>% base::subset(Kingdom == "k__Archaea" | Kingdom == "k__Bacteria")
meco.fun$tax_table %>% base::subset(Kingdom == "k__Fungi")

meco.bac$filter_pollution(taxa = c("mitochondria", "chloroplast"))
meco.fun$filter_pollution(taxa = c("mitochondria", "chloroplast"))

meco.bac$tidy_dataset()                                                         
meco.fun$tidy_dataset()

## Convert back to ps
ps.meco.bac = meco2phyloseq(meco.bac)
ps.meco.fun = meco2phyloseq(meco.fun)

ps.bac
ps.meco.bac
ps.fun
ps.meco.fun

## Normalize
library(microbiomeMarker) # TSS normalization
library(dplyr)            # data wrangling

ps.bac.tss = ps.meco.bac %>% microbiomeMarker::normalize(method = "TSS")
ps.fun.tss = ps.meco.fun %>% microbiomeMarker::normalize(method = "TSS")
```

### 2 \| Isolate top n taxa

First, we need to extract the req. df from the phyloseq object.

```{r isolate, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Load required packages
library(phyloseq)
library(ggplot2)
library(dplyr)
library(ggpubr)

## Aggregate to taxonomic level of choice 
ps.bac.tss_class <- tax_glom(ps.bac.tss, taxrank = "Class")
ps.fun.tss_order <- tax_glom(ps.fun.tss, taxrank = "Order")

## Extract abundance data as a df /w psmelt (& combine with metadata & taxonomy)
df.bac.tss_class <- psmelt(ps.bac.tss_class) %>% filter(Sample != "DSTB")
df.fun.tss_order <- psmelt(ps.fun.tss_order)

## Isolate top 10-15 taxa per site based on mean abundance
taxa_TF_bac.class <- df.bac.tss_class %>% 
  filter(site == "TF", Sample != "DSTB") %>%
  group_by(Class) %>% summarise(mean_abund = mean(Abundance)) %>% 
  top_n(15, mean_abund) %>%
  pull(Class)
taxa_WSE_bac.class <- df.bac.tss_class %>% 
  filter(site == "WSE") %>%
  group_by(Class) %>% 
  summarise(mean_abund = mean(Abundance)) %>% 
  top_n(15, mean_abund) %>%
  pull(Class)

taxa_TF_fun.order <- df.fun.tss_order %>% 
  filter(site == "TF") %>%
  group_by(Order) %>% 
  summarise(mean_abund = mean(Abundance)) %>% 
  top_n(15, mean_abund) %>%
  pull(Order)
taxa_WSE_fun.order <- df.fun.tss_order %>% 
  filter(site == "WSE") %>%
  group_by(Order) %>% 
  summarise(mean_abund = mean(Abundance)) %>% 
  top_n(15, mean_abund) %>%
  pull(Order)

## Subset og data to only include the top taxa
df_top_taxa_TF_bac.class <- df.bac.tss_class %>% 
  filter(site == "TF", Sample != "DSTB", Class %in% taxa_TF_bac.class)
df_top_taxa_WSE_bac.class <- df.bac.tss_class %>% 
  filter(site == "WSE", Class %in% taxa_WSE_bac.class)

df_top_taxa_TF_fun.order <- df.fun.tss_order %>% 
  filter(site == "TF", Order %in% taxa_TF_fun.order)
df_top_taxa_WSE_fun.order <- df.fun.tss_order %>% 
  filter(site == "WSE", Order %in% taxa_WSE_fun.order)
```

Now, we can finally plot the mean abundances per forest type.

```{r plot, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(ggpubr)

## BACteria Forest type 1
pvals1 <- compare_means(Abundance ~ season,
                        group.by = "Class",
                        data = df_top_taxa_TF_bac.class, 
                        method = "wilcox.test")
y.pos1 <- df_top_taxa_TF_bac.class %>% 
  mutate(Class = gsub("^c__", "", Class)) %>%
  group_by(Class) %>% 
  summarise(y.position = max(Abundance, na.rm = TRUE) *1)

pvals1 <- pvals1 %>% 
  mutate(Class = gsub("^c__", "", Class)) %>% 
  left_join(y.pos1, by = "Class") %>% 
  dplyr::distinct(Class, .keep_all = TRUE)

df_top_taxa_TF_bac.class %>%
  mutate(Class = gsub("^c__", "", Class)) %>%
  ggplot(aes(x = reorder(Class, -Abundance), y = Abundance, fill = season)) +
  geom_bar(stat = "summary", fun = "mean", position = position_dodge()) +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  coord_cartesian(ylim = c(0,1.0))+
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.8)) +
  scale_fill_manual(values = c("lightgreen","seagreen3")) +
  stat_pvalue_manual(pvals1, 
                     label = "p.signif", 
                     x = "Class", 
                     y.position = "y.position", 
                     tip.length = 0)+
  stat_summary(fun.data = mean_sd, 
               geom = "errorbar",  
               aes(ymin = after_stat(y)),
               position = position_dodge(width = 0.8), 
               width = 0.25, linewidth = 0.6) 

## BACteria Forest type 2
pvals2 <- compare_means(Abundance ~ season, 
                        group.by = "Class",
                        data = df_top_taxa_WSE_bac.class,
                        method = "wilcox.test")
y.pos2 <- df_top_taxa_WSE_bac.class %>% 
  mutate(Class = gsub("^c__", "", Class)) %>%
  group_by(Class) %>% 
  summarise(y.position = max(Abundance, na.rm = TRUE) *1)

pvals2 <- pvals2 %>% 
  mutate(Class = gsub("^c__", "", Class)) %>% 
  left_join(y.pos2, by = "Class") %>% 
  dplyr::distinct(Class, .keep_all = TRUE)

df_top_taxa_WSE_bac.class %>%
  mutate(Class = gsub("^c__", "", Class)) %>%
  ggplot(aes(x = reorder(Class, -Abundance), y = Abundance, fill = season)) +
  geom_bar(stat = "summary", fun = "mean", position = position_dodge()) +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  coord_cartesian(ylim = c(0,0.8))+
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.8)) +
  scale_fill_manual(values = c("bisque2","bisque4")) +
  stat_pvalue_manual(pvals2, 
                     label = "p.signif", 
                     x = "Class", 
                     y.position = "y.position", 
                     tip.length = 0)+
  stat_summary(fun.data = mean_sd, 
               geom = "errorbar",  
               aes(ymin = after_stat(y)),
               position = position_dodge(width = 0.8), 
               width = 0.25, linewidth = 0.6) 

## FUNgi Forest type 1
pvals3 <- compare_means(Abundance ~ season, 
                        group.by = "Order",
                        data = df_top_taxa_TF_fun.order, 
                        method = "wilcox.test")
y.pos3 <- df_top_taxa_TF_fun.order %>% 
  mutate(Order = gsub("^o__", "", Order)) %>%
  group_by(Order) %>% 
  summarise(y.position = max(Abundance, na.rm = TRUE) *1)

pvals3 <- pvals3 %>% 
  mutate(Order = gsub("^o__", "", Order)) %>% 
  left_join(y.pos3, by = "Order") %>% 
  dplyr::distinct(Order, .keep_all = TRUE)

df_top_taxa_TF_fun.order %>%
  mutate(Order = gsub("^o__", "", Order)) %>%
  ggplot(aes(x = reorder(Order, -Abundance), y = Abundance, fill = season)) +
  geom_bar(stat = "summary", fun = "mean", position = position_dodge()) +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  coord_cartesian(ylim = c(0,0.8))+
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.8)) +
  scale_fill_manual(values = c("lightgreen","seagreen3")) +
  stat_pvalue_manual(pvals3, 
                     label = "p.signif", 
                     x = "Order", 
                     y.position = "y.position", 
                     tip.length = 0)+
  stat_summary(fun.data = mean_sd, 
               geom = "errorbar", 
               aes(ymin = after_stat(y)), 
               position = position_dodge(width = 0.8), 
               width = 0.25, linewidth = 0.6) 

## FUNgi Forest type 2
pvals4 <- compare_means(Abundance ~ season, 
                        group.by = "Order",
                        data = df_top_taxa_WSE_fun.order, 
                        method = "wilcox.test")
y.pos4 <- df_top_taxa_WSE_fun.order %>% 
  mutate(Order = gsub("^o__", "", Order)) %>%
  group_by(Order) %>% 
  summarise(y.position = max(Abundance, na.rm = TRUE) *1)

pvals4 <- pvals4 %>% 
  mutate(Order = gsub("^o__", "", Order)) %>% 
  left_join(y.pos4, by = "Order") %>% 
  dplyr::distinct(Order, .keep_all = TRUE)

df_top_taxa_WSE_fun.order %>%
  mutate(Order = gsub("^o__", "", Order)) %>%
  ggplot(aes(x = reorder(Order, -Abundance), y = Abundance, fill = season)) +
  geom_bar(stat = "summary", fun = "mean", position = position_dodge()) +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  coord_cartesian(ylim = c(0,0.6))+
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.8)) +
  scale_fill_manual(values = c("bisque2","bisque4")) +
  stat_pvalue_manual(pvals4, 
                     label = "p.signif", 
                     x = "Order", 
                     y.position = "y.position", 
                     tip.length = 0)+
  stat_summary(fun.data = mean_sd, 
               geom = "errorbar", 
               aes(ymin = after_stat(y)),
               position = position_dodge(width = 0.8), 
               width = 0.25, linewidth = 0.6) 
```

### **3 \| OPTIONAL: Create zoomed-in panels for taxa w/ low-abundance visibility**

```{r plot2, tidy=FALSE, message=FALSE, warning=FALSE, results='hide', echo=FALSE}
## Define regular plot as object
gg_top15_BAC.TF = df_top_taxa_TF_bac.class %>%
  mutate(Class = gsub("^c__", "", Class)) %>%
  ggplot(aes(x = reorder(Class, -Abundance), y = Abundance, fill = season)) +
  geom_bar(stat = "summary", fun = "mean", position = position_dodge()) +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  coord_cartesian(ylim = c(0,1.0))+
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.8)) +
  scale_fill_manual(values = c("lightgreen","seagreen3")) +
  stat_pvalue_manual(pvals1, 
                     label = "p.signif", 
                     x = "Class", 
                     y.position = "y.position", 
                     tip.length = 0)+
  stat_summary(fun.data = mean_sd, 
               geom = "errorbar",  
               aes(ymin = after_stat(y)),
               position = position_dodge(width = 0.8), 
               width = 0.25, linewidth = 0.6) 

gg_top15_BAC.WS = df_top_taxa_WSE_bac.class %>%
  mutate(Class = gsub("^c__", "", Class)) %>%
  ggplot(aes(x = reorder(Class, -Abundance), y = Abundance, fill = season)) +
  geom_bar(stat = "summary", fun = "mean", position = position_dodge()) +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  coord_cartesian(ylim = c(0,0.8))+
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.8)) +
  scale_fill_manual(values = c("bisque2","bisque4")) +
  stat_pvalue_manual(pvals2, 
                     label = "p.signif", 
                     x = "Class", 
                     y.position = "y.position", 
                     tip.length = 0)+
  stat_summary(fun.data = mean_sd, 
               geom = "errorbar",  
               aes(ymin = after_stat(y)),
               position = position_dodge(width = 0.8), 
               width = 0.25, linewidth = 0.6) 

gg_top15_FUN.TF = df_top_taxa_TF_fun.order %>%
  mutate(Order = gsub("^o__", "", Order)) %>%
  ggplot(aes(x = reorder(Order, -Abundance), y = Abundance, fill = season)) +
  geom_bar(stat = "summary", fun = "mean", position = position_dodge()) +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  coord_cartesian(ylim = c(0,0.8))+
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.8)) +
  scale_fill_manual(values = c("lightgreen","seagreen3")) +
  stat_pvalue_manual(pvals3, 
                     label = "p.signif", 
                     x = "Order", 
                     y.position = "y.position", 
                     tip.length = 0)+
  stat_summary(fun.data = mean_sd, 
               geom = "errorbar", 
               aes(ymin = after_stat(y)), 
               position = position_dodge(width = 0.8), 
               width = 0.25, linewidth = 0.6) 

gg_top15_FUN.WS = df_top_taxa_WSE_fun.order %>%
  mutate(Order = gsub("^o__", "", Order)) %>%
  ggplot(aes(x = reorder(Order, -Abundance), y = Abundance, fill = season)) +
  geom_bar(stat = "summary", fun = "mean", position = position_dodge()) +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  coord_cartesian(ylim = c(0,0.6))+
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.8)) +
  scale_fill_manual(values = c("bisque2","bisque4")) +
  stat_pvalue_manual(pvals4, 
                     label = "p.signif", 
                     x = "Order", 
                     y.position = "y.position", 
                     tip.length = 0)+
  stat_summary(fun.data = mean_sd, 
               geom = "errorbar", 
               aes(ymin = after_stat(y)),
               position = position_dodge(width = 0.8), 
               width = 0.25, linewidth = 0.6) 
```

```{r plot3, tidy=FALSE, message=FALSE, warning=FALSE, results='hide', fig.show='hide'}
library(dplyr)
library(cowplot)

## Create zoomed-in plot 
gg_top15_BAC.TF_zoom = gg_top15_BAC.TF +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100), 
                     breaks = c(0.0,0.01,0.02,0.03,0.04,0.05)) +
  theme(axis.text.x = element_blank(), 
        axis.title.y = element_blank(), 
        legend.position = "none") +
  scale_fill_manual(values = c("lightgreen","seagreen3")) +
  coord_cartesian(ylim = c(0,0.05))

gg_top15_BAC.WS_zoom = gg_top15_BAC.WS +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100), 
                     breaks = c(0.0,.005,0.01,0.015,0.02)) +
  theme(axis.text.x = element_blank(), 
        axis.title.y = element_blank(), 
        legend.position = "none") +
  scale_fill_manual(values = c("bisque2","bisque4")) +
  coord_cartesian(ylim = c(0,0.02))

gg_top15_FUN.TF_zoom = gg_top15_FUN.TF +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100), 
                     breaks = c(0.0,0.01,0.02,0.03,0.04,0.05,0.06)) +
  theme(axis.text.x = element_blank(), 
        axis.title.y = element_blank(), 
        legend.position = "none") +
  scale_fill_manual(values = c("lightgreen","seagreen3")) +
  coord_cartesian(ylim = c(0,0.06))

gg_top15_FUN.WS_zoom = gg_top15_FUN.WS +
  labs(x = NULL, y = "Mean %RA") +
  ggpubr::theme_pubr() +
  scale_y_continuous(labels = function(x) sprintf("%.1f%%", x * 100), 
                     breaks = c(0.0,0.02,0.04,0.05,0.06,0.07)) +
  theme(axis.text.x = element_blank(), 
        axis.title.y = element_blank(), 
        legend.position = "none") +
  scale_fill_manual(values = c("bisque2","bisque4")) +
  coord_cartesian(ylim = c(0,0.07))

## Plot og plot w/ zoomed-in panel
ggdraw() +
  draw_plot(gg_top15_BAC.TF) +
  draw_plot(gg_top15_BAC.TF_zoom, 
            x = 0.2, y = 0.55, width = 0.75, height = 0.3) 
ggdraw() +
  draw_plot(gg_top15_BAC.WS) +
  draw_plot(gg_top15_BAC.WS_zoom, 
            x = 0.2, y = 0.55, width = 0.75, height = 0.3) 
ggdraw() +
  draw_plot(gg_top15_FUN.TF) +
  draw_plot(gg_top15_FUN.TF_zoom, 
            x = 0.2, y = 0.55, width = 0.75, height = 0.3) 
ggdraw() +
  draw_plot(gg_top15_FUN.WS) +
  draw_plot(gg_top15_FUN.WS_zoom, 
            x = 0.2, y = 0.55, width = 0.75, height = 0.3) 
```
