---
title: "Fig. S4: Differentially abundant pathways associated with nutrient cycling or drought response."
author: "Jessica Finck"
date: "2026-02-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

This script used the respective KEGG orthologue (KO) table to extract predicted functions associated with nutrient cycling and drought response. Using the applied KO filter, it then summarizes the cumulative functional potential for these two life history strategies, and runs statistical tests to infer statistical significance.

### Pre-processing of predicted functional data

First, we import the data into R. KOs are filtered based on a separate .xlsx file, which is based on publications from [**Xue et al.**](https://doi.org/10.3389/fmicb.2021.698286), [**Martins et al.**](https://doi.org/10.1016/j.apsoil.2024.105715) and [**Nalven et al.**](https://doi.org/10.1111/1462-2920.15121) Please see the corresponding file in the [**corresponding github**](https://github.com/JFinck1994/Finck_2026_TropMiCom_under_drought).

**Important!** Here, we filter out 1 specific sample (DTSB) due to likely contamination.

**Functional table (KO_per_sample.tsv):**

| function | smp1 | smp2 | ... |
|----------|------|------|-----|
| K00001   |      |      |     |
| K00002   |      |      |     |
| ...      |      |      |     |

**Metadata table (BAC_silva138.2.xlsx \> meta_all):**

| smp_id   | site | season | ... |
|----------|------|--------|-----|
| sample 1 | TF   | wet    |     |
| sample 2 | WS   | dry    |     |
| ...      |      |        |     |

```{r import, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Load data 
library(readxl)
library(dplyr)
library(tidyr)
library(tibble)

bac_dat <- read_xlsx("BAC_silva138.2.xlsx", sheet = "meta_all") %>% 
  as.data.frame() %>% 
  filter(smp_id != "DSTB")
bac_KO = read.delim(
  "KO_per_sample.tsv", 
  row.names = 1, 
  check.names = FALSE) %>% 
  dplyr::select(-"DSTB")

## Normalize data
## ! Important: Rows must be samples
library(janitor)
library(vegan)

bac_KO = as.data.frame(x = t(bac_KO), stringsAsFactors = FALSE)
bac_KO = decostand(bac_KO, method = "total")

## check if it was normalized correctly (rows == 1, columns =! 1)
rowSums(bac_KO)
colSums(bac_KO)

## Define subset data based on forest type
tf_samples  <- bac_dat$smp_id[bac_dat$site == "TF"] %>% as.list()
ws_samples <- bac_dat$smp_id[bac_dat$site == "WSE"]

## Subset data 
bac_KO_tf <- bac_KO[row.names(bac_KO) %in% tf_samples,] 
bac_KO_ws <- bac_KO[row.names(bac_KO) %in% ws_samples,]

bac_dat_tf <- bac_dat[bac_dat$smp_id %in% tf_samples,]
bac_dat_ws <- bac_dat[bac_dat$smp_id %in% ws_samples,]

## Load KO filter 
library(readxl)
KO_core_genes <- read_excel("KO_core_genes.XLSX", sheet = "combined_genes_DOM")
```

### Calculation of predicted functional z-scores

We calculate z-scores using base R's `scale()` function.

```{r format, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Define K_number col
k_number.tf <- grep("^K", colnames(bac_KO_tf), value = TRUE)
k_number.ws <- grep("^K", colnames(bac_KO_ws), value = TRUE)

## Calculate z-score 
z_KO.tf <- bac_KO_tf
z_KO.tf[k_number.tf] <- scale(bac_KO_tf[k_number.tf]) 

z_KO.ws <- bac_KO_ws
z_KO.ws[k_number.ws] <- scale(bac_KO_ws[k_number.ws]) 

## Check if z-scores calc correctly 
colMeans(z_KO.tf[k_number.tf], na.rm = TRUE)     # should be ~0
apply(z_KO.tf[k_number.tf], 2, sd, na.rm = TRUE) # should be ~1

colMeans(z_KO.ws[k_number.ws], na.rm = TRUE)     # should be ~0
apply(z_KO.ws[k_number.ws], 2, sd, na.rm = TRUE) # should be ~1

## Join z-scores w/ metadata 
y_levels <- c("TF-wet","WS-wet","TF-dry","WS-dry")

z_KO.tf.long <- z_KO.tf %>%
  rownames_to_column("smp_id") %>% 
  pivot_longer(cols = -smp_id, names_to = "k_number", values_to = "z_score") %>%
  semi_join(KO_core_genes, by = "k_number") %>%
  left_join(bac_dat_tf %>% select(smp_id, season, site), by = "smp_id") %>%
  left_join(KO_core_genes %>% 
              select(k_number, bar_filter, bar_group, bar_pathway, z_group1, z_group2, z_group3), by = "k_number") %>%
  mutate(site_season = paste(site, season, sep = "-"), site_season = factor(site_season, levels = y_levels))

z_KO.ws.long <- z_KO.ws %>%
  rownames_to_column("smp_id") %>% 
  pivot_longer(cols = -smp_id, names_to = "k_number", values_to = "z_score") %>%
  semi_join(KO_core_genes, by = "k_number") %>%
  left_join(bac_dat_ws %>% select(smp_id, season, site), by = "smp_id") %>%
  mutate(site = recode(site, "WSE" = "WS")) %>%
  left_join(KO_core_genes %>% 
              select(k_number, bar_filter, bar_group, bar_pathway, z_group1, z_group2, z_group3), by = "k_number") %>%
  mutate(site_season = paste(site, season, sep = "-"), site_season = factor(site_season, levels = y_levels))

## Aggregate z-scores for heatmap
z_KO.tf.agg <- z_KO.tf.long %>%
  group_by(z_group3, site_season, z_group1, z_group2) %>%  
  summarise(mean_z = mean(z_score, na.rm = TRUE), .groups = "drop") %>%
  mutate(sig_flag = ifelse(mean_z <= -1.96 | mean_z >= 1.96, "*", NA_character_))

z_KO.ws.agg <- z_KO.ws.long %>%
  group_by(z_group3, site_season, z_group1, z_group2) %>%  
  summarise(mean_z = mean(z_score, na.rm = TRUE), .groups = "drop") %>%
  mutate(sig_flag = ifelse(mean_z <= -1.96 | mean_z >= 1.96, "*", NA_character_))

## Bind z-scores for separate forest types into 1 df
z_KO.all.agg <- bind_rows(z_KO.tf.agg, z_KO.ws.agg)

## Export z-score
library(readr)
write_csv(z_KO.all.agg, "z-scores.csv")
```

### Plotting z-score heatmap

Re-import z-scores together with metadata for easier heatmap plotting, using `pheatmap`.

```{r heatmap, message=FALSE, warning=FALSE, results='hide'}
library(pheatmap)
library(tidyverse)
library(readxl)

## Load data
db_pheat <- read_excel("pheatmap_data_example.xlsx", sheet = "avg_values") 
db_pheat <- db_pheat %>%
  column_to_rownames(var = names(db_pheat)[1]) %>% 
  as.matrix() 

## Load metadata annotation
ann_db <- read_excel("pheatmap_data_example.xlsx", sheet = "annotations") 
ann_db <- ann_db %>%
  column_to_rownames(var = names(ann_db)[1]) 

## Load pw annotation
enz_groups_db <- read_excel("pheatmap_data_example.xlsx", sheet = "annotations2") 
enz_groups_db <-  enz_groups_db %>%
  column_to_rownames(var = names(enz_groups_db)[1]) %>% 
  select(-z_group1) %>% 
  rename("Strategy" = z_group1_alt) %>% 
  rename("Function" = z_group2) 

## define annotation colors 
ann_colors <- list(
  Function = c(
    "Carbon cycling" =  "#77ff05",
    "Nitrogen cycling" = "#5bbe01",
    "Sulfur cycling" = "#3e8201",
    "Phosphorous cycling" = "#316501", 
    "ABC transporters" = "chocolate",
    "Osmoprotectant biosynthesis" = "chocolate1",
    "Sporulation" = "chocolate2",
    "Universal stress regulator" = "chocolate3",
    "DNA protection and repair" = "chocolate4"), 
  Strategy = c("Nutrient cycling" = "#69de02",
               "Drought response" = "orange1"),
  Site = c("TF-dry" = "lightgreen",
           "TF-wet" = "lightgreen",
           "WS-dry" = "bisque2",
           "WS-wet" = "bisque2"),
  Season = c("dry" = "firebrick", 
             "wet" = "royalblue"))

## heatmap
pheatmap(db_pheat,
         cluster_rows = T, cluster_cols = T,
         scale = "row",
         clustering_method = 'ward.D2',
         colorRampPalette(c("navy", "white", "seagreen3"))(12), 
         border_color = "gray75",
         cutree_rows = 4, cutree_cols = 4,
         angle_col = 45,
         annotation_row = enz_groups_db, 
         annotation_col = ann_db,
         annotation_colors = ann_colors)
```
